---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# RaJIVE

<!-- badges: start -->
<!-- badges: end -->

RaJIVE (Robust Angle based Joint and Individual Variation Explained) is a robust alternative to the aJIVE method for the estimation of joint and individual components in the presence of outliers in multi-source data. It decomposes the multi-source data into joint, individual and residual (noise) contributions. The decomposition is robust with respect to outliers and other types of noises present in the data.

## Installation

You can install the released version of RaJIVE from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("RaJIVE")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ericaponzi/RaJIVE")
```
## Example

This is a basic example which shows how to use RaJIVE on simple simulated data: 


### Running robust aJIVE


```{r example}
library(RaJIVE)
## basic example code
n <- 50
pks <- c(100, 80, 50)
Y <- ajive.data.sim(K =3, rankJ = 3, rankA = c(7, 6, 4), n = n,
                   pks = pks, dist.type = 1)

initial_signal_ranks <-  c(7, 6, 4)
data.ajive <- list((Y$sim_data[[1]]), (Y$sim_data[[2]]), (Y$sim_data[[3]]))
ajive.results.robust <- Rajive(data.ajive, initial_signal_ranks)

```


The function returns a list of class `"rajive"` containing the RaJIVE decomposition, with the joint component (shared across data sources), individual component (data source specific) and residual component for each data source. 



### Inspecting the decomposition

- Print a concise overview:

```{r print_summary}
print(ajive.results.robust)
```

- Summary table of all ranks:

```{r all_ranks}
summary(ajive.results.robust)
get_all_ranks(ajive.results.robust)
```

- Joint rank:

```{r ranks}
get_joint_rank(ajive.results.robust)
```

- Individual ranks:

```{r}
get_individual_rank(ajive.results.robust, 1)
get_individual_rank(ajive.results.robust, 2)
get_individual_rank(ajive.results.robust, 3)
```

- Shared joint scores (n × joint_rank matrix):

```{r joint_scores, eval = FALSE}
get_joint_scores(ajive.results.robust)
```

- Block-specific scores and loadings:

```{r}
# Joint scores for block 1
get_block_scores(ajive.results.robust, k = 1, type = "joint")

# Individual loadings for block 2
get_block_loadings(ajive.results.robust, k = 2, type = "individual")
```

- Full reconstructed matrices (J, I, or E) for a block:

```{r block_matrix, eval = FALSE}
J1 <- get_block_matrix(ajive.results.robust, k = 1, type = "joint")
I2 <- get_block_matrix(ajive.results.robust, k = 2, type = "individual")
E3 <- get_block_matrix(ajive.results.robust, k = 3, type = "noise")
```

### Visualizing results

- Heatmap decomposition:

```{r}
decomposition_heatmaps_robustH(data.ajive, ajive.results.robust)
```

- Proportion of variance explained (as a list):

```{r}
showVarExplained_robust(ajive.results.robust, data.ajive)
```

- Proportion of variance explained (as a bar chart):

```{r var_plot, eval = FALSE}
plot_variance_explained(ajive.results.robust, data.ajive)
```

- Scatter plot of scores (e.g. joint component 1 vs 2 for block 1):

```{r scores_plot, eval = FALSE}
plot_scores(ajive.results.robust, k = 1, type = "joint",
            comp_x = 1, comp_y = 2)

# Colour points by a grouping variable
group_labels <- rep(c("A", "B"), each = n / 2)
plot_scores(ajive.results.robust, k = 1, type = "joint",
            comp_x = 1, comp_y = 2, group = group_labels)
```

### Jackstraw significance testing

After running the RaJIVE decomposition, you can test which variables in each data block have statistically significantly non-zero joint loadings using the jackstraw permutation test:

```{r jackstraw, eval = FALSE}
# Run jackstraw test (increase n_null to 50-100 for publication-quality results)
js <- jackstraw_rajive(ajive.results.robust, data.ajive,
                       alpha = 0.05, n_null = 10,
                       correction = "bonferroni")

# Print a concise summary table
print(js)

# Get a data frame summary
summary(js)
```

- Retrieve significant variables for a given block and component:

```{r sig_vars, eval = FALSE}
get_significant_vars(js, block = 1, component = 1)
```

- Visualize jackstraw results (three plot types available):

```{r plot_jackstraw, eval = FALSE}
# P-value histogram
plot_jackstraw(js, type = "pvalue_hist", block = 1, component = 1)

# F-statistic vs -log10(p-value) scatter plot
plot_jackstraw(js, type = "scatter", block = 1, component = 1)

# Heatmap of -log10(p-value) across all joint components for one block
plot_jackstraw(js, type = "loadings_significance", block = 1)
```

## Function reference

### Core decomposition

| Function | Description |
|---|---|
| `Rajive()` | Run the RaJIVE decomposition on a list of data matrices. Returns an object of class `"rajive"`. |
| `ajive.data.sim()` | Simulate multi-block data with known joint and individual structure for testing and benchmarking. |

### Rank accessors

| Function | Description |
|---|---|
| `get_joint_rank()` | Extract the estimated joint rank from a `"rajive"` object. |
| `get_individual_rank()` | Extract the individual rank for a specific data block. |
| `get_all_ranks()` | Return a `data.frame` of joint and individual ranks for all blocks at once. |

### Component accessors

| Function | Description |
|---|---|
| `get_joint_scores()` | Return the shared \eqn{n \times \text{joint\_rank}} joint score matrix. |
| `get_block_scores()` | Return the score matrix (U) for a given block and component type (joint or individual). |
| `get_block_loadings()` | Return the loading matrix (V) for a given block and component type. |
| `get_block_matrix()` | Return the full reconstructed matrix (J, I, or E) for a given block and component type. |

### S3 methods for `"rajive"` objects

| Function | Description |
|---|---|
| `print.rajive()` | Print a concise summary of ranks for a `"rajive"` object. |
| `summary.rajive()` | Return and print a `data.frame` of all estimated ranks. |

### Variance explained

| Function | Description |
|---|---|
| `showVarExplained_robust()` | Compute the proportion of variance explained by joint, individual, and residual components for each block (returns a list). |
| `plot_variance_explained()` | Stacked bar chart of variance explained by each component and block. |

### Visualisation

| Function | Description |
|---|---|
| `decomposition_heatmaps_robustH()` | Heatmaps of the raw data and the joint, individual, and noise components for all blocks. |
| `plot_scores()` | Scatter plot of two score components for a given block (joint or individual), with optional group colouring. |

### Jackstraw significance testing

| Function | Description |
|---|---|
| `jackstraw_rajive()` | Run the jackstraw permutation test to identify variables with significantly non-zero joint loadings. Returns a `"jackstraw_rajive"` object. |
| `print.jackstraw_rajive()` | Print a significance table for a `"jackstraw_rajive"` object. |
| `summary.jackstraw_rajive()` | Return and print a `data.frame` summary of jackstraw results. |
| `get_significant_vars()` | Extract significant variable names/indices for a given block and component from jackstraw results. |
| `plot_jackstraw()` | Diagnostic plots for jackstraw results: p-value histogram, F-stat scatter plot, or loadings significance heatmap. |

