---
title: "Where the Time Goes: rajiveplus vs RaJIVE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Where the Time Goes: rajiveplus vs RaJIVE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
run_heavy <- FALSE         # set to TRUE once to regenerate .rds cache files
data_dir  <- "data"        # relative to vignettes/
```




## The question

Matrix decompositions ask a simple question and pay for it many times: which
directions in these blocks are shared? In Robust aJIVE, the expensive part is
the repeated robust SVD. `rajiveplus` moves that bottleneck into C++ through
RcppArmadillo, while the original
[RaJIVE](https://github.com/ericaponzi/RaJIVE) package keeps the computation in
pure R.

The useful benchmark is therefore not abstract. It is: when does that C++
implementation change the working scale of an analysis? We look along five
axes:

| Benchmark | Varies | Fixed |
|-----------|--------|-------|
| Head-to-head timing | implementation | K=2, n=100, p=500, joint_rank=2 |
| Scaling vs n | n in {50,100,200} | p=500 |
| Scaling vs p | p in {100,500,1000,2000} | n=100 |
| Parallelism (rajiveplus only) | num_cores in {1,2,4} | n=200, p=1000 |
| Peak memory | implementation | K=2, n=100, p=500 |

Two details keep the comparison honest. First, every benchmark fixes
`joint_rank = 2`, so we time the decomposition rather than stochastic rank
selection. Second, `RaJIVE` registers two parallel workers internally, so the
head-to-head runs give `rajiveplus` the same two-core budget.

This is a timing vignette, not an interpretation vignette. Helpers such as
`extract_components()`, `plot_components()`, `associate_components()`, and
`assess_stability()` appear in the application vignettes where their outputs
matter scientifically.

---

## Setup

Install the reference implementation if it is not already available:

```{r}
if (!requireNamespace("RaJIVE", quietly = TRUE))
  remotes::install_github("ericaponzi/RaJIVE")
```




Both packages export a function named `Rajive()`. The benchmark code therefore
uses explicit namespaces whenever the two implementations are compared.

```{r}
library(rajiveplus)
library(microbenchmark)
library(bench)
library(peakRAM)
library(ggplot2)
```




We also register one two-worker `doParallel` cluster before benchmarking. That
keeps worker start-up from masquerading as algorithmic cost.

```{r}
doParallel::registerDoParallel(2L)
```




The main benchmark data set has two blocks, 100 samples, and 500 features per
block:

```{r}
set.seed(42)
Y      <- ajive.data.sim(K = 2, rankJ = 2, rankA = c(5, 5),
                         n = 100, pks = c(500, 500))
blocks <- Y$sim_data
```




---

## First look

Before the formal summaries, a single run gives the scale of the difference.

```{r}
t_ru <- system.time(
  rajiveplus::Rajive(blocks, c(5, 5), joint_rank = 2L, full = FALSE, num_cores = 2L)
)
t_rj <- system.time(
  RaJIVE::Rajive(blocks, c(5, 5), joint_rank = 2L, full = FALSE)
)
cat("rajiveplus:", round(t_ru["elapsed"], 2), "s\n")
cat("RaJIVE:     ", round(t_rj["elapsed"], 2), "s\n")
cat("Speedup:    ", round(t_rj["elapsed"] / t_ru["elapsed"], 1), "x\n")
```




---

## Repeated timing

Five repeated runs are enough to show the timing distribution without turning
the vignette into a compute job.

```{r}
bm_mb <- microbenchmark(
  rajiveplus = rajiveplus::Rajive(blocks, c(5, 5),
                                   joint_rank = 2L, full = FALSE, num_cores = 2L),
  RaJIVE      = RaJIVE::Rajive(blocks, c(5, 5), joint_rank = 2L, full = FALSE),
  times       = 5L
)
saveRDS(bm_mb, file.path(data_dir, "bench_results.rds"))
```

```{r}
bm_mb <- readRDS(file.path(data_dir, "bench_results.rds"))
print(bm_mb, unit = "s", order = "median")
```

```{r}
autoplot(bm_mb) +
  labs(title = "microbenchmark: rajiveplus vs RaJIVE",
       subtitle = "K=2, n=100, p=500, joint_rank=2, num_cores=2, 5 reps",
       x = "Time (ms)", y = NULL) +
  theme_bw()
```




---

## bench::mark comparison

`bench::mark` adds medians and spread. Its memory column measures R-level
allocations only, so native Armadillo allocations are addressed later with
`peakRAM`.

```{r}
bm_bench <- bench::mark(
  rajiveplus = rajiveplus::Rajive(blocks, c(5, 5),
                                   joint_rank = 2L, full = FALSE, num_cores = 2L),
  RaJIVE      = RaJIVE::Rajive(blocks, c(5, 5), joint_rank = 2L, full = FALSE),
  iterations  = 3L,
  check       = FALSE,
  memory      = FALSE
)
saveRDS(bm_bench, file.path(data_dir, "bench_results_bench.rds"))
```

```{r}
bm_bench <- readRDS(file.path(data_dir, "bench_results_bench.rds"))
knitr::kable(
  summary(bm_bench)[, c("expression", "min", "median", "mem_alloc", "n_itr")],
  caption = "bench::mark results (3 iterations)"
)
```

```{r}
plot(bm_bench) +
  labs(title = "bench::mark: rajiveplus vs RaJIVE",
       subtitle = "R-heap allocations only; see peakRAM for total RSS") +
  theme_bw()
```




---

## Scaling with samples and features

```{r}
ns <- c(50L, 100L, 200L)

scaling_n_rows <- lapply(ns, function(n_i) {
  set.seed(42)
  Y_i    <- ajive.data.sim(K = 2, rankJ = 2, rankA = c(5, 5),
                           n = n_i, pks = c(500, 500))
  blks_i <- Y_i$sim_data
  t_ru   <- system.time(rajiveplus::Rajive(blks_i, c(5, 5),
                                            joint_rank = 2L, full = FALSE, num_cores = 2L))["elapsed"]
  t_rj   <- system.time(RaJIVE::Rajive(blks_i, c(5, 5), joint_rank = 2L, full = FALSE))["elapsed"]
  data.frame(n = n_i, rajiveplus = t_ru, RaJIVE = t_rj)
})
scaling_n_df <- do.call(rbind, scaling_n_rows)
```

```{r}
ps <- c(100L, 500L, 1000L, 2000L)

scaling_p_rows <- lapply(ps, function(p_i) {
  set.seed(42)
  Y_i    <- ajive.data.sim(K = 2, rankJ = 2, rankA = c(5, 5),
                           n = 100, pks = c(p_i, p_i))
  blks_i <- Y_i$sim_data
  t_ru   <- system.time(rajiveplus::Rajive(blks_i, c(5, 5),
                                            joint_rank = 2L, full = FALSE, num_cores = 2L))["elapsed"]
  t_rj   <- system.time(RaJIVE::Rajive(blks_i, c(5, 5), joint_rank = 2L, full = FALSE))["elapsed"]
  data.frame(p = p_i, rajiveplus = t_ru, RaJIVE = t_rj)
})
scaling_p_df <- do.call(rbind, scaling_p_rows)

saveRDS(list(vs_n = scaling_n_df, vs_p = scaling_p_df),
        file.path(data_dir, "scaling_results.rds"))
```

```{r}
scaling      <- readRDS(file.path(data_dir, "scaling_results.rds"))
scaling_n_df <- scaling$vs_n
scaling_p_df <- scaling$vs_p
```

```{r}
df_n_long <- reshape(scaling_n_df,
                     varying   = c("rajiveplus", "RaJIVE"),
                     v.names   = "elapsed",
                     timevar   = "package",
                     times     = c("rajiveplus", "RaJIVE"),
                     direction = "long")

ggplot(df_n_long, aes(x = n, y = elapsed, colour = package, group = package)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  scale_colour_manual(values = c(rajiveplus = "#2166ac", RaJIVE = "#d6604d")) +
  labs(title = "Elapsed time vs sample size (n)",
       subtitle = "K=2, p=500, joint_rank=2, num_cores=2",
       x = "Sample size (n)", y = "Elapsed time (s)", colour = NULL) +
  theme_bw()
```

```{r}
df_p_long <- reshape(scaling_p_df,
                     varying   = c("rajiveplus", "RaJIVE"),
                     v.names   = "elapsed",
                     timevar   = "package",
                     times     = c("rajiveplus", "RaJIVE"),
                     direction = "long")

ggplot(df_p_long, aes(x = p, y = elapsed, colour = package, group = package)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  scale_x_log10() +
  scale_colour_manual(values = c(rajiveplus = "#2166ac", RaJIVE = "#d6604d")) +
  labs(title = "Elapsed time vs feature count (p)",
       subtitle = "K=2, n=100, joint_rank=2, num_cores=2 (log x-axis)",
       x = "Features per block (p)", y = "Elapsed time (s)", colour = NULL) +
  theme_bw()
```

```{r}
speedup_df <- data.frame(
  p       = scaling_p_df$p,
  speedup = round(scaling_p_df$RaJIVE / scaling_p_df$rajiveplus, 1)
)
knitr::kable(speedup_df,
             col.names = c("p (features per block)", "Speedup (x)"),
             caption   = "rajiveplus speedup over RaJIVE by feature count")
```




---

## What do extra cores buy?

`rajiveplus` can spread work across cores through `parallel::mclapply`. Here
we ask whether that matters on a larger example with 200 samples and 1000
features per block.

```{r}
set.seed(42)
Y_par    <- ajive.data.sim(K = 2, rankJ = 2, rankA = c(5, 5),
                           n = 200, pks = c(1000, 1000))
blks_par <- Y_par$sim_data

parallel_rows <- lapply(c(1L, 2L, 4L), function(nc) {
  t_i <- system.time(
    rajiveplus::Rajive(blks_par, c(5, 5), joint_rank = 2L, full = FALSE, num_cores = nc)
  )["elapsed"]
  data.frame(num_cores = nc, elapsed = t_i)
})
parallel_df         <- do.call(rbind, parallel_rows)
parallel_df$speedup <- parallel_df$elapsed[1] / parallel_df$elapsed

saveRDS(parallel_df, file.path(data_dir, "parallel_results.rds"))
```

```{r}
parallel_df <- readRDS(file.path(data_dir, "parallel_results.rds"))
knitr::kable(parallel_df, digits = 2,
             col.names = c("num_cores", "Elapsed (s)", "Speedup (x)"),
             caption   = "rajiveplus parallelism scaling (n=200, p=1000)")
```

```{r}
p1 <- ggplot(parallel_df, aes(x = num_cores, y = elapsed)) +
  geom_line(colour = "#2166ac", linewidth = 1) +
  geom_point(size = 2.5, colour = "#2166ac") +
  labs(title = "Elapsed time vs num_cores",
       x = "num_cores", y = "Elapsed time (s)") +
  theme_bw()

p2 <- ggplot(parallel_df, aes(x = num_cores, y = speedup)) +
  geom_line(colour = "#1b7837", linewidth = 1) +
  geom_point(size = 2.5, colour = "#1b7837") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  labs(title = "Speedup vs num_cores",
       subtitle = "Dashed = ideal linear speedup",
       x = "num_cores", y = "Speedup (x)") +
  theme_bw()

if (requireNamespace("patchwork", quietly = TRUE)) {
  print(patchwork::wrap_plots(p1, p2, ncol = 2))
} else {
  print(p1); print(p2)
}
```




---

## Memory footprint

Speed is only half the story. The C++ RobustSVD allocates through Armadillo, so
R's heap counters miss part of the footprint. `peakRAM` watches the process at
the operating-system level and gives the fairer comparison.

```{r}
mem_ru <- peakRAM::peakRAM(
  rajiveplus::Rajive(blocks, c(5, 5), joint_rank = 2L, full = FALSE, num_cores = 2L)
)
mem_rj <- peakRAM::peakRAM(
  RaJIVE::Rajive(blocks, c(5, 5), joint_rank = 2L, full = FALSE)
)
peakram_df <- data.frame(
  package     = c("rajiveplus", "RaJIVE"),
  peak_ram_mb = c(mem_ru$Peak_RAM_Used_MiB, mem_rj$Peak_RAM_Used_MiB),
  elapsed_s   = c(mem_ru$Elapsed_Time_sec,  mem_rj$Elapsed_Time_sec)
)
saveRDS(peakram_df, file.path(data_dir, "peakram_single.rds"))
```

```{r}
peakram_df <- readRDS(file.path(data_dir, "peakram_single.rds"))
knitr::kable(peakram_df,
             col.names = c("Package", "Peak RAM (MiB)", "Elapsed (s)"),
             digits    = 1,
             caption   = "Peak RSS memory (peakRAM); includes C++ heap allocations")
```

```{r}
ggplot(peakram_df, aes(x = package, y = peak_ram_mb, fill = package)) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c(rajiveplus = "#2166ac", RaJIVE = "#d6604d")) +
  labs(title  = "Peak RAM usage (peakRAM)",
       subtitle = "K=2, n=100, p=500, joint_rank=2, num_cores=2",
       x = NULL, y = "Peak RAM (MiB)") +
  theme_bw() + theme(legend.position = "none")
```




> **Why peakRAM?** `bench::mark` reports R-heap allocation. The native heap used
> by Armadillo and LAPACK is outside that accounting. `peakRAM` captures the
> process-level resident set size, which is the memory pressure the scheduler
> sees.

---

## Session Info

```{r}
sessionInfo()
```
