---
title: "A Lightweight Runtime Smoke Test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Lightweight Runtime Smoke Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
data_dir <- "data"
```

## The question

The package vignette is deliberately light. It checks that the current
`rajiveplus` runtime path can be compared with a locally installed copy of the
original `RaJIVE` package, but it does not run the full benchmark grid.

The full benchmark artifact lives in `inst/benchmarks/benchmarking_heavy.Rmd`
and is rendered only by SLURM after this package-vignette suite succeeds. That
heavy artifact regenerates the cache files used for release notes and
performance inspection:

```{r cache-status}
cache_files <- c(
  "bench_results.rds",
  "bench_results_bench.rds",
  "scaling_results.rds",
  "parallel_results.rds",
  "peakram_single.rds"
)
cache_paths <- file.path(data_dir, cache_files)
cache_status <- data.frame(
  file = cache_files,
  present = file.exists(cache_paths),
  modified = ifelse(file.exists(cache_paths),
                    format(file.info(cache_paths)$mtime, "%Y-%m-%d %H:%M"),
                    NA_character_),
  stringsAsFactors = FALSE
)
knitr::kable(cache_status, caption = "Heavy benchmark cache status")
```

## Setup

This vignette does not install packages from the network. The SLURM render node
must already have the local benchmark dependencies available.

```{r setup}
required <- c("rajiveplus", "RaJIVE", "ggplot2")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0L) {
  stop(
    "Missing local package(s) required for the benchmark smoke test: ",
    paste(missing, collapse = ", "),
    ". Install them in the R4_51 environment before rendering this vignette.",
    call. = FALSE
  )
}

library(rajiveplus)
library(ggplot2)
```

Both packages export a function named `Rajive()`, so the comparison uses
explicit namespaces. The `rajiveplus` call uses `identifiability_norm = "l1"`
only for parity with the original implementation's post-selection filter.

## Tiny smoke benchmark

The data set is intentionally small: two blocks, 40 samples, and 60 features
per block. This is not a performance claim; it is a fast integration check that
both implementations can run on the current SLURM software stack.

```{r smoke-data}
set.seed(42)
Y <- ajive.data.sim(K = 2, rankJ = 1L, rankA = c(2L, 2L),
                    n = 40L, pks = c(60L, 60L), dist.type = 1)
blocks <- Y$sim_data
```

```{r smoke-run}
elapsed <- function(expr) {
  unname(system.time(force(expr))[["elapsed"]])
}

t_ru <- elapsed(
  rajiveplus::Rajive(
    blocks, c(2L, 2L),
    joint_rank = 1L,
    full = FALSE,
    n_wedin_samples = NA,
    n_rand_dir_samples = NA,
    num_cores = 1L,
    seed = 42L,
    identifiability_norm = "l1"
  )
)

t_rj <- elapsed(
  RaJIVE::Rajive(blocks, c(2L, 2L), joint_rank = 1L, full = FALSE)
)

timings <- data.frame(
  package = c("rajiveplus", "RaJIVE"),
  elapsed_s = c(t_ru, t_rj),
  stringsAsFactors = FALSE
)
knitr::kable(timings, digits = 3, caption = "Tiny smoke benchmark")
```

```{r smoke-plot}
ggplot(timings, aes(x = package, y = elapsed_s, fill = package)) +
  geom_col(width = 0.55) +
  scale_fill_manual(values = c(rajiveplus = "#2166ac", RaJIVE = "#d6604d")) +
  labs(title = "Tiny smoke benchmark",
       subtitle = "K=2, n=40, p=60, joint_rank=1",
       x = NULL, y = "Elapsed time (s)") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r smoke-ratio}
cat("Smoke-test speedup:", round(t_rj / t_ru, 1), "x\n")
```

## Heavy benchmark handoff

The heavy SLURM render runs the repeated timings, scaling sweeps, parallel
scaling, and peak-memory comparison. It should be submitted with an `afterok`
dependency on the light vignette array so expensive benchmarking starts only
after this smoke test and the other package vignettes are green.

```{r session-info}
sessionInfo()
```
