---
title: "What Is Shared Across CLL Omics Layers?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What Is Shared Across CLL Omics Layers?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
run_heavy <- FALSE  # set to FALSE after .rds files are generated
data_dir  <- "data"
```



## The question

The CLL cohort asks a familiar multi-omics question: when transcript abundance,
DNA methylation, and drug response are measured on overlapping patients, which
axes of variation are genuinely shared? This vignette applies `rajiveplus` to
the public Chronic Lymphocytic Leukaemia cohort from Dietrich et al. (2018)
([PMID 29227291](https://pubmed.ncbi.nlm.nih.gov/29227291/)), distributed
through `BloodCancerMultiOmics2017`. The same data were used in the original
MOFA2 vignette (Argelaguet et al. 2020), giving a useful point of comparison.

We use three approximately Gaussian views:

| Block | Source object | Biological meaning |
|-------|---------------|--------------------|
| mRNA | `dds` (DESeqDataSet, 63 677 genes × 136 samples) | Transcriptional programs |
| Methylation | `methData` (RangedSummarizedExperiment, 5 000 CpGs × 196 samples) | Epigenetic state |
| Drug viability | `lpdAll` (ExpressionSet, 448 drug–dose features × 249 samples) | Pharmacological sensitivity |

> **Why no mutations block?** The MOFA2 tutorial includes somatic mutations as
> a fourth block using a Bernoulli likelihood. RaJIVE inherits the JIVE
> Gaussian assumption, so binary mutation matrices are not used as input here.
> IGHV mutation status and trisomy 12 instead enter as metadata for
> interpretation and clinical validation.

The three-way sample overlap after restricting to CLL patients is **~121 samples**.

The analysis follows five questions:

1. Which joint sources of variation are shared across all three layers?
2. Which feature weights make those axes biologically legible?
3. Do joint factors align with IGHV status or trisomy 12?
4. Which features are significant under `jackstraw_rajive()`?
5. Can joint factors predict missing IGHV status and time to treatment?

---

## Prerequisites

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("BloodCancerMultiOmics2017")
install.packages(c("survival", "survminer", "maxstat", "randomForest",
                   "pheatmap", "ggrepel", "patchwork"))
```

```{r libs}
library(rajiveplus)
library(ggplot2)
library(pheatmap)
```



---

## Data loading

```{r load-data, eval = run_heavy}
library(BloodCancerMultiOmics2017)
data("dds")        # DESeqDataSet: RNA-seq, 63677 genes x 136 samples
data("methData")   # RangedSummarizedExperiment: 5000 CpGs x 196 samples
data("lpdAll")     # ExpressionSet: 448 viab features + metadata x 249 samples
data("patmeta")    # Clinical metadata data.frame, rownames = patient H-IDs

cat("Raw dimensions:\n")
cat("  RNA-seq (dds):", dim(dds), "\n")
cat("  Methylation (methData):", dim(methData), "\n")
cat("  lpdAll (all features):", dim(lpdAll), "\n")
cat("  patmeta:", dim(patmeta), "\n")
```

```{r preprocess, eval = run_heavy}
# ---- RNA-seq ----
# dds column names are integers; map to patient H-IDs via colData$PatID
rna_hids    <- SummarizedExperiment::colData(dds)[["PatID"]]
counts_mat  <- SummarizedExperiment::assay(dds)       # genes x samples
colnames(counts_mat) <- rna_hids

# log2(count + 1) normalisation; transpose to samples x genes
mrna_full   <- t(log2(counts_mat + 1))                # samples x genes

# Select top 500 highly variable genes
rv_mrna     <- apply(mrna_full, 2, var, na.rm = TRUE)
mrna_top    <- mrna_full[, order(rv_mrna, decreasing = TRUE)[1:500], drop = FALSE]

# ---- Methylation ----
# methData: features x samples; transpose to samples x CpGs
meth_full   <- t(SummarizedExperiment::assay(methData))  # samples x CpGs
rv_meth     <- apply(meth_full, 2, var, na.rm = TRUE)
meth_top    <- meth_full[, order(rv_meth, decreasing = TRUE)[1:500], drop = FALSE]

# ---- Drug viability ----
# lpdAll: filter to 'viab' feature type, then transpose
viab_idx    <- which(Biobase::fData(lpdAll)[["type"]] == "viab")
viab_full   <- t(Biobase::exprs(lpdAll)[viab_idx, ])    # samples x drug-doses
rv_viab     <- apply(viab_full, 2, var, na.rm = TRUE)
viab_top    <- viab_full[, order(rv_viab, decreasing = TRUE)[1:min(448, ncol(viab_full))],
                          drop = FALSE]

# ---- Three-way sample overlap ----
common3 <- Reduce(intersect, list(
  rownames(mrna_top),
  rownames(meth_top),
  rownames(viab_top)
))
# Restrict to CLL diagnosis
cll_flag  <- patmeta[common3, "Diagnosis"] == "CLL"
cll_flag[is.na(cll_flag)] <- FALSE
common_cll <- common3[cll_flag]
cat("Three-way overlap (CLL patients):", length(common_cll), "samples\n")

# ---- Subset and scale ----
scale_block <- function(m) {
  m_s <- scale(m, center = TRUE, scale = TRUE)
  m_s[is.nan(m_s)] <- 0
  m_s
}

blocks <- list(
  mRNA        = scale_block(mrna_top[common_cll, , drop = FALSE]),
  Methylation = scale_block(meth_top[common_cll, , drop = FALSE]),
  Drugs       = scale_block(viab_top[common_cll, , drop = FALSE])
)
meta_cc <- patmeta[common_cll, , drop = FALSE]

cat("Final block dimensions:\n")
print(lapply(blocks, dim))

saveRDS(list(blocks = blocks, meta = meta_cc),
        file.path(data_dir, "cll_preprocessed.rds"))
```

```{r load-preprocessed}
pp     <- readRDS(file.path(data_dir, "cll_preprocessed.rds"))
blocks <- pp$blocks
meta   <- pp$meta
cat("Samples:", nrow(blocks[[1]]), "\n")
cat("Block dimensions:\n"); print(lapply(blocks, dim))
```



---

## How much signal should each block contribute?

```{r scree-compute, eval = run_heavy}
svd_list <- lapply(blocks, function(m) svd(m, nu = 0, nv = 0)$d)
saveRDS(svd_list, file.path(data_dir, "cll_svd_list.rds"))
```

```{r scree-load}
svd_list <- readRDS(file.path(data_dir, "cll_svd_list.rds"))
```

```{r scree-plot}
block_names <- c("mRNA", "Methylation", "Drugs")
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (i in seq_along(svd_list)) {
  sv  <- svd_list[[i]]
  pve <- sv^2 / sum(sv^2) * 100
  plot(seq_along(pve[1:30]), pve[1:30], type = "b", pch = 19,
       xlab = "Component", ylab = "% Variance Explained",
       main = block_names[i], col = "#2166ac")
  abline(v = c(15L, 10L, 10L)[i], col = "#d6604d", lty = 2)
}
par(mfrow = c(1, 1))
```

The dashed red lines mark deliberately conservative initial signal ranks:
**mRNA = 15**, **Methylation = 10**, and **Drugs = 10**. They define the
candidate subspaces from which RaJIVE will estimate the shared structure.

---

## Fit RaJIVE

```{r run-rajive, eval = run_heavy}
set.seed(42)
cll_fit <- Rajive(
  blocks               = blocks,
  initial_signal_ranks = c(15L, 10L, 10L),
  full                 = TRUE,
  n_wedin_samples      = 100L,
  n_rand_dir_samples   = 100L,
  joint_rank           = NA,
  num_cores            = 12L
)
saveRDS(cll_fit, file.path(data_dir, "cll_rajive_results.rds"))
```

```{r load-rajive}
cll_fit <- readRDS(file.path(data_dir, "cll_rajive_results.rds"))
cat("Estimated joint rank:", get_joint_rank(cll_fit), "\n")
print(get_all_ranks(cll_fit))
```

---

## What did the decomposition find?

Before interpreting any factor, inspect the skeleton of the decomposition: the
joint rank and the block-specific individual ranks. `summarize_components()`
returns that structure as a compact table.

```{r comp-structure}
comp_tbl <- summarize_components(ajive_output = cll_fit,
                                 summary_type = "components")
knitr::kable(comp_tbl,
             caption = "Decomposition structure: joint and individual ranks per block")
```

The pairwise score plot is the RaJIVE analogue of a factor-correlation check.
RaJIVE imposes orthogonality within the joint subspace, so strong residual
correlation would be a warning sign rather than a biological discovery.

```{r factor-pairs}
plot_components(cll_fit, plot_type = "pairs", type = "joint") +
  labs(title    = "Pairwise joint factor scores",
       subtitle = "Orthogonality check — factors should be uncorrelated")
```

---

## Was the rank decision convincing?

AJIVE selects the joint rank by asking whether observed joint singular values
clear two data-adaptive thresholds: the Wedin perturbation bound and the
random-direction bound.

```{r rank-diag-table}
diag_long <- extract_components(cll_fit, what = "rank_diagnostics",
                                format = "long")
knitr::kable(head(diag_long, 10),
             caption = "Top AJIVE rank-diagnostic rows")
```

The composite diagnostic plot puts the evidence side by side: bound
distributions, per-block singular values, and the final threshold decision.

```{r rank-diag-composite, fig.height = 6}
plot_components(cll_fit, plot_type = "ajive_diagnostic") +
  labs(title = "CLL AJIVE full diagnostic")
```

```{r rank-threshold-plot}
plot_components(cll_fit, plot_type = "rank_threshold") +
  labs(title = "CLL AJIVE rank-threshold diagnostics")
```

```{r diag-summary}
diag_tbl <- summarize_components(ajive_output = cll_fit,
                                 summary_type = "diagnostics")
knitr::kable(diag_tbl, digits = 4,
             caption = "AJIVE rank-selection diagnostics summary")
```

---

## Which clinical labels follow the joint factors?

Next we test whether joint scores line up with clinical variables. These
p-values are exploratory because the score-estimation uncertainty is not
propagated in this section.

```{r assoc-ighv}
assoc_ighv <- associate_components(
  ajive_output = cll_fit,
  metadata     = meta,
  variable     = "IGHV",
  mode         = "categorical",
  type         = "joint"
)
knitr::kable(assoc_ighv, digits = 4,
             caption = "Association: joint factors vs IGHV mutation status")
```

```{r assoc-trisomy}
assoc_t12 <- associate_components(
  ajive_output = cll_fit,
  metadata     = meta,
  variable     = "trisomy12",
  mode         = "categorical",
  type         = "joint"
)
knitr::kable(assoc_t12, digits = 4,
             caption = "Association: joint factors vs trisomy 12 status")
```

```{r assoc-plot}
# associate_components() already sets a 'variable' column; just rbind
assoc_combined <- rbind(assoc_ighv, assoc_t12)
plot_components(assoc_results = assoc_combined,
                plot_type     = "association") +
  labs(title = "Factor-covariate association summary")
```

### Does the rank survive resampling?

```{r stability-run, eval = run_heavy}
stab_rank <- assess_stability(
  ajive_output          = cll_fit,
  blocks                = blocks,
  initial_signal_ranks  = c(15L, 10L, 10L),
  target                = "joint_rank",
  B                     = 20L,
  sample_frac           = 0.8,
  num_cores             = 2L
)
saveRDS(stab_rank, file.path(data_dir, "cll_stability.rds"))
```

```{r stability-table}
stab_rank <- readRDS(file.path(data_dir, "cll_stability.rds"))
knitr::kable(as.data.frame(stab_rank$rank_table),
             col.names = c("Joint rank", "Frequency"),
             caption   = "Bootstrap joint-rank stability (B = 20)")
```

---

## Where is the variation?

The variance decomposition is the high-level map of the fit. It shows whether
each omics layer is dominated by shared joint structure, layer-specific
individual structure, or residual variation.

```{r var-plot}
plot_variance_explained(cll_fit, blocks) +
  labs(title    = "CLL multi-omics: variance decomposition",
       subtitle = "Joint, Individual, and Noise components per block") +
  theme_bw()
```

```{r var-table}
ve_tbl <- summarize_components(ajive_output = cll_fit,
                               blocks       = blocks,
                               summary_type = "variance")
knitr::kable(ve_tbl, digits = 2,
             caption = "Variance explained by component type per block (%)")
```

---

## What do the joint factors mean?

### Factor 1 vs Factor 2: IGHV status

```{r scores-setup}
joint_sc <- as.data.frame(get_joint_scores(cll_fit))
colnames(joint_sc) <- paste0("Factor", seq_len(ncol(joint_sc)))
rownames(joint_sc) <- rownames(blocks[[1]])

common_ids <- intersect(rownames(joint_sc), rownames(meta))
joint_sc_m <- joint_sc[common_ids, , drop = FALSE]
meta_m     <- meta[common_ids, , drop = FALSE]
```

```{r scores-f1f2}
nf <- min(2, ncol(joint_sc_m))
plot_df <- data.frame(joint_sc_m[, 1:nf, drop = FALSE],
                      IGHV = factor(meta_m$IGHV))
colnames(plot_df)[1:2] <- c("Factor1", "Factor2")

ggplot(plot_df, aes(x = Factor1, y = Factor2, colour = IGHV)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_colour_manual(values = c("M" = "#2166ac", "U" = "#d6604d"),
                      na.value = "grey70") +
  labs(title    = "CLL joint scores: Factor 1 vs Factor 2",
       subtitle = "Coloured by IGHV mutation status (M = mutated, U = unmutated)",
       colour   = "IGHV") +
  theme_bw()
```

### Factor 1 vs Factor 3: IGHV status and trisomy 12

This plot mirrors the central MOFA2 CLL display: a low-dimensional map of
patients and molecular subgroups. Factor 1 is checked for the IGHV mutated (M)
versus unmutated (U) split; Factor 3, when present, is checked for trisomy 12.

```{r scores-f1f3}
if (ncol(joint_sc_m) >= 3) {
  plot_df2 <- data.frame(
    Factor1   = joint_sc_m[["Factor1"]],
    Factor3   = joint_sc_m[["Factor3"]],
    IGHV      = factor(meta_m$IGHV),
    Trisomy12 = factor(meta_m$trisomy12)
  )

  ggplot(plot_df2,
         aes(x = Factor1, y = Factor3, colour = IGHV, shape = Trisomy12)) +
    geom_point(size = 2.5, alpha = 0.85) +
    scale_colour_manual(values = c("M" = "#2166ac", "U" = "#d6604d"),
                        na.value = "grey70") +
    scale_shape_manual(values = c("0" = 16, "1" = 17), na.value = 4) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title    = "CLL joint scores: Factor 1 vs Factor 3",
         subtitle = "Colour = IGHV status; shape = trisomy 12 (1 = present)",
         colour   = "IGHV", shape = "Trisomy12") +
    theme_bw()
} else {
  message("Joint rank < 3; Factor 1 vs Factor 3 plot skipped.")
}
```

### Factor–covariate correlation heatmap

```{r factor-cov-heatmap}
covariates <- intersect(c("IGHV", "treatedAfter", "died", "trisomy12"),
                        colnames(meta_m))
cov_vals   <- meta_m[, covariates, drop = FALSE]
n_factors  <- ncol(joint_sc_m)

cor_mat <- matrix(NA, nrow = n_factors, ncol = ncol(cov_vals),
                  dimnames = list(colnames(joint_sc_m), covariates))
for (fi in seq_len(n_factors)) {
  for (ci in seq_along(covariates)) {
    cv <- suppressWarnings(as.numeric(cov_vals[[ci]]))
    if (sum(!is.na(cv)) > 5)
      cor_mat[fi, ci] <- cor(joint_sc_m[[fi]], cv, use = "complete.obs")
  }
}

pheatmap(cor_mat,
         cluster_rows = FALSE, cluster_cols = FALSE,
         color    = colorRampPalette(c("#d6604d", "white", "#2166ac"))(100),
         breaks   = seq(-1, 1, length.out = 101),
         main     = "Factor-covariate correlations (Pearson r)",
         fontsize = 11)
```

---

## Which features give the factors their names?

### Top feature loadings (bar chart)

`plot_components(plot_type = "top_features")` gives a signed loading chart,
the RaJIVE counterpart of MOFA2's top-weight view. Features near zero do not
define the axis; features with large absolute loadings do.

```{r top-features-bar}
plot_components(cll_fit,
                plot_type = "top_features",
                type      = "joint",
                block     = 1L,
                component = 1L,
                top_n     = 20L) +
  labs(title    = "Top 20 mRNA loadings — Factor 1 (joint)",
       subtitle = "Positive = higher in IGHV-mutated; negative = higher in IGHV-unmutated")
```

### Factor value vs gene expression scatter

The scatter plots connect a latent score back to measured expression. For each
top-weighted gene, Factor 1 is shown against normalized expression and colored
by IGHV status, so the sign of the loading can be read biologically.

```{r data-scatter, fig.height = 4}
# Retrieve top positive and top negative loading genes
top_pos <- rank_features(ajive_output = cll_fit, mode = "top_loadings",
                         type = "joint", block = 1L, component = 1L,
                         n = 8L, signed = TRUE)
top_pos <- head(top_pos[top_pos$loading > 0, ], 4)

top_neg <- rank_features(ajive_output = cll_fit, mode = "top_loadings",
                         type = "joint", block = 1L, component = 1L,
                         n = 8L, signed = TRUE)
top_neg <- head(top_neg[top_neg$loading < 0, ], 4)

mrna_mat  <- blocks[["mRNA"]]          # scaled; samples x genes
f1_scores <- joint_sc_m[["Factor1"]]

make_scatter <- function(genes, title_suffix) {
  genes <- intersect(genes, colnames(mrna_mat))
  if (length(genes) == 0L) return(NULL)
  df <- do.call(rbind, lapply(genes, function(g) {
    data.frame(Factor1    = f1_scores,
               Expression = mrna_mat[names(f1_scores), g],
               Gene       = g,
               IGHV       = factor(meta_m[names(f1_scores), "IGHV"]),
               stringsAsFactors = FALSE)
  }))
  ggplot(df, aes(x = Factor1, y = Expression, colour = IGHV)) +
    geom_point(size = 1.5, alpha = 0.7) +
    facet_wrap(~Gene, scales = "free_y", nrow = 1) +
    scale_colour_manual(values = c("M" = "#2166ac", "U" = "#d6604d"),
                        na.value = "grey70") +
    labs(title  = paste("Factor 1 vs mRNA expression:", title_suffix),
         x      = "Factor 1 score",
         y      = "Scaled expression",
         colour = "IGHV") +
    theme_bw(base_size = 10)
}

p_pos <- make_scatter(top_pos$feature_name, "top positive weights")
p_neg <- make_scatter(top_neg$feature_name, "top negative weights")

if (!is.null(p_pos)) print(p_pos)
if (!is.null(p_neg)) print(p_neg)
```

---

## Which feature weights are statistically supported?

```{r jackstraw-run, eval = run_heavy}
set.seed(42)
cll_js <- jackstraw_rajive(
  ajive_output = cll_fit,
  blocks       = blocks,
  alpha        = 0.05,
  n_null       = 50L,
  correction   = "BH"
)
saveRDS(cll_js, file.path(data_dir, "cll_jackstraw_results.rds"))
```

```{r jackstraw-load}
cll_js <- readRDS(file.path(data_dir, "cll_jackstraw_results.rds"))
print(cll_js)
```

`summarize_components()` turns the jackstraw result into counts of significant
features for each block-component pair at the chosen FDR threshold.

```{r jackstraw-sig-summary}
sig_tbl <- summarize_components(jackstraw_result = cll_js,
                                summary_type     = "significance")
knitr::kable(sig_tbl, digits = 4,
             caption = "Significant features per block/component (BH FDR 5%)")
```

```{r jackstraw-pvalue-hist}
plot_jackstraw(cll_js, type = "pvalue_hist", block = 1L, component = 1L) +
  labs(title    = "Jackstraw: mRNA block, Factor 1",
       subtitle = "Observed vs null F-statistic distributions") +
  theme_bw()
```

A volcano plot shows the gene-level tradeoff between effect size and
significance.

```{r jackstraw-volcano}
plot_components(jackstraw_result = cll_js,
                plot_type        = "volcano",
                block            = 1L,
                component        = 1L) +
  labs(title = "Volcano: mRNA block, Factor 1 (joint)")
```

```{r jackstraw-sig-genes}
sig_mrna <- get_significant_vars(cll_js, block = 1L, component = 1L)
cat("Significant mRNA features (Factor 1, BH alpha = 0.05):", nrow(sig_mrna), "\n")
if (nrow(sig_mrna) > 0)
  knitr::kable(head(sig_mrna, 20),
               caption = "Top significant mRNA features in Factor 1")
```

---

## Molecular signatures

```{r top-loadings-table}
top25_tbl <- rank_features(
  ajive_output = cll_fit,
  mode         = "top_loadings",
  type         = "joint",
  block        = 1L,
  component    = 1L,
  n            = 25L,
  signed       = TRUE
)
knitr::kable(head(top25_tbl, 10), digits = 4,
             caption = "Top-ranked mRNA loadings from rank_features()")
```

The component heatmap sorts samples by the Factor 1 score and displays the
top-loaded features as rows. It is the RaJIVE counterpart of MOFA2's
data-heatmap view.

```{r component-heatmap}
plot_components(cll_fit,
                plot_type = "component_heatmap",
                type      = "joint",
                block     = 1L,
                component = 1L) +
  labs(title = "mRNA joint component heatmap — Factor 1")
```

---

## Can factors predict clinical labels?

### IGHV status (Random Forest)

We train a random forest on the joint factor scores to predict IGHV status.
Samples missing IGHV labels can then be imputed, echoing the prediction step in
the MOFA2 CLL vignette.

```{r rf-ighv}
if (requireNamespace("randomForest", quietly = TRUE)) {
  rf_df <- data.frame(
    joint_sc_m,
    IGHV = factor(meta_m$IGHV[match(rownames(joint_sc_m), rownames(meta_m))])
  )
  rf_df_train <- rf_df[!is.na(rf_df$IGHV), , drop = FALSE]
  set.seed(42)
  rf_fit <- randomForest::randomForest(IGHV ~ ., data = rf_df_train, ntree = 500L)
  cat("OOB error (IGHV prediction):",
      round(rf_fit$err.rate[nrow(rf_fit$err.rate), "OOB"] * 100, 1), "%\n")
  print(rf_fit$confusion)

  # Impute all samples (including those with missing labels)
  rf_df$IGHV_pred   <- stats::predict(rf_fit, rf_df)
  rf_df$IGHV_source <- ifelse(
    is.na(meta_m$IGHV[match(rownames(rf_df), rownames(meta_m))]),
    "Predicted", "Observed"
  )
}
```

The plot places observed and imputed IGHV labels back into Factor 1 versus
Factor 3 space, making the imputation visually auditable.

```{r rf-plot}
if (exists("rf_df") && ncol(joint_sc_m) >= 3) {
  pred_df <- data.frame(
    Factor1     = joint_sc_m[rownames(rf_df), "Factor1"],
    Factor3     = joint_sc_m[rownames(rf_df), "Factor3"],
    IGHV_pred   = rf_df$IGHV_pred,
    IGHV_source = rf_df$IGHV_source
  )
  ggplot(pred_df,
         aes(x = Factor1, y = Factor3,
             colour = IGHV_pred, shape = IGHV_source)) +
    geom_point(size = 2.5, alpha = 0.85) +
    scale_colour_manual(values = c("M" = "#2166ac", "U" = "#d6604d")) +
    scale_shape_manual(values  = c("Observed" = 16, "Predicted" = 17)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(title  = "IGHV prediction: observed vs imputed labels",
         colour = "IGHV (predicted)",
         shape  = "Label source") +
    theme_bw()
}
```

---

## Survival Analysis

### Data preparation

```{r surv-setup}
if (requireNamespace("survival", quietly = TRUE)) {
  # T5 = time to first treatment (years); treatedAfter = event indicator
  surv_df <- data.frame(
    joint_sc_m,
    TTT          = meta_m$T5[match(rownames(joint_sc_m), rownames(meta_m))],
    treatedAfter = as.integer(
      meta_m$treatedAfter[match(rownames(joint_sc_m), rownames(meta_m))]
    )
  )
  surv_df <- surv_df[!is.na(surv_df$TTT) & !is.na(surv_df$treatedAfter), ,
                     drop = FALSE]
  cat("Samples with survival data:", nrow(surv_df), "\n")
}
```

### Cox proportional-hazards model

```{r cox-model}
if (requireNamespace("survival", quietly = TRUE) &&
    exists("surv_df") && nrow(surv_df) > 10) {
  factor_cols <- paste(paste0("Factor", seq_len(get_joint_rank(cll_fit))),
                       collapse = " + ")
  cox_fit <- survival::coxph(
    as.formula(paste("survival::Surv(TTT, treatedAfter) ~", factor_cols)),
    data = surv_df
  )
  knitr::kable(round(summary(cox_fit)$coefficients, 4),
               caption = "Cox PH: joint factors vs time to treatment")
}
```

### Hazard ratio forest plot

```{r hr-forest}
if (requireNamespace("survival", quietly = TRUE) && exists("cox_fit")) {
  s     <- summary(cox_fit)
  coefs <- s[["coefficients"]]
  ci    <- s[["conf.int"]]

  hr_df <- data.frame(
    factor = factor(rownames(coefs), levels = rev(rownames(coefs))),
    p      = coefs[, "Pr(>|z|)"],
    HR     = ci[, "exp(coef)"],
    lower  = ci[, "lower .95"],
    upper  = ci[, "upper .95"]
  )

  ggplot(hr_df, aes(x = factor, y = HR, ymin = lower, ymax = upper)) +
    geom_pointrange(colour = "#619CFF") +
    geom_hline(aes(yintercept = 1), linetype = "dotted") +
    coord_flip() +
    labs(title = "Hazard ratios: joint factors vs time to treatment",
         x = "", y = "Hazard Ratio (95% CI)") +
    theme_bw()
}
```

### Kaplan-Meier curve (optimal cutpoint)

Instead of splitting at the median, we use the maximally selected rank
statistic (`surv_cutpoint()` from **survminer**) to choose the Factor 1
threshold. This matches the approach used in the MOFA2 CLL vignette.

```{r km-plot}
if (requireNamespace("survminer", quietly = TRUE) &&
    requireNamespace("survival",  quietly = TRUE) &&
    exists("surv_df") && nrow(surv_df) > 10) {

  # Optimal cutpoint via maximally selected rank statistic
  cut_res <- survminer::surv_cutpoint(
    data      = surv_df,
    time      = "TTT",
    event     = "treatedAfter",
    variables = "Factor1"
  )
  opt_cut <- cut_res$cutpoint$cutpoint
  surv_df$Factor1_group <- ifelse(surv_df$Factor1 > opt_cut, "High", "Low")

  km_fit <- survival::survfit(
    survival::Surv(TTT, treatedAfter) ~ Factor1_group, data = surv_df
  )
  survminer::ggsurvplot(
    km_fit, data = surv_df,
    pval        = TRUE,
    conf.int    = TRUE,
    risk.table  = TRUE,
    palette     = c("#d6604d", "#2166ac"),
    title       = sprintf("Time to treatment by Factor 1 (cutpoint = %.2f)", opt_cut),
    xlab        = "Time (years)",
    ylab        = "Survival probability (%)",
    fun         = function(y) y * 100,
    legend.labs = c("High Factor 1", "Low Factor 1")
  )
}
```

---

## How does this compare with MOFA2?

The MOFA2 CLL analysis (Argelaguet et al. 2020) reported four latent factors.
The table below maps the main biological interpretations across methods.
Factor numbers are not directly comparable: MOFA2 orders factors by total
variance explained, whereas RaJIVE factors reflect the geometry of the joint
subspace.

```{r mofa2-comparison}
jrank   <- get_joint_rank(cll_fit)
signals <- c(
  "IGHV status / B-cell differentiation (cf. MOFA2 Factor 1)",
  "Trisomy 12 / proliferative signalling (cf. MOFA2 Factor 3)",
  rep("Inspect loadings — no MOFA2 analogue identified", max(0L, jrank - 2L))
)[seq_len(jrank)]
evidence <- c(
  "High |r| with IGHV; separates M/U in score plots",
  "Inspect mRNA + drug loadings; check trisomy12 association",
  rep("See loading heatmap", max(0L, jrank - 2L))
)[seq_len(jrank)]

comp_df <- data.frame(
  `RaJIVE Factor`     = paste0("Factor ", seq_len(jrank)),
  `Biological signal` = signals,
  `Key evidence`      = evidence,
  check.names = FALSE
)
knitr::kable(comp_df,
             caption = "RaJIVE-MOFA2 factor correspondence (preliminary)")
```

> **Methodological note.** MOFA2 is a probabilistic latent factor model with
> view-specific likelihoods; RaJIVE is a deterministic matrix decomposition
> that maximizes explained variance under a Gaussian assumption. Both methods
> identify IGHV status as a dominant CLL axis. The differences matter:
> RaJIVE gives a hard joint/individual split, its jackstraw test gives
> feature-level significance without a sparse prior, and MOFA2 natively
> handles non-Gaussian views such as binary mutations and counts.

---

## Session Info

<details>
<summary>**Session Info**</summary>

```{r session-info}
sessionInfo()
```

</details>
