---
title: "A Guided Tour of the rajiveplus API"
author: "rajiveplus package authors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{A Guided Tour of the rajiveplus API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
run_heavy <- FALSE   # set to TRUE once to regenerate all cached .rds files
data_dir  <- "data"
```


## The question

What can you ask of a fitted RaJIVE model? This gallery walks through the
exported user-facing functions in **rajiveplus** using one small three-block
simulation. The example is deliberately light so the whole tour can run quickly;
for a biological analysis, see `vignette("cll_application")`.

Each stop follows the same rhythm: fit or extract one object, inspect the
output, then say what scientific question the output can answer.

### Roadmap

| Category | Functions |
|---|---|
| Simulation | `ajive.data.sim` |
| Core decomposition | `Rajive` |
| S3 — *rajive* class | `print`, `summary` |
| Rank accessors | `get_joint_rank`, `get_individual_rank`, `get_all_ranks` |
| Component extractors | `get_joint_scores`, `get_block_scores`, `get_block_loadings`, `get_block_matrix` |
| Diagnostic extraction | `extract_components` (5 `what` modes) |
| Decomposition visualization | `data_heatmap`, `decomposition_heatmaps_robustH` |
| Variance explained | `showVarExplained_robust`, `plot_variance_explained` |
| Score plots | `plot_scores`, `plot_components` (`pairs`, `density`) |
| AJIVE diagnostic plots | `plot_components` (`rank_threshold`, `bound_distributions`, `ajive_diagnostic`) |
| Feature visualization | `plot_components` (`top_features`, `component_heatmap`) |
| S3 autoplot / fortify | `autoplot`, `fortify` |
| Feature ranking | `rank_features` (4 modes) + 3 wrappers |
| Association testing | `associate_components` (4 modes) + 3 wrappers; `plot_components` (`association`) |
| Stability | `assess_stability` (3 targets) + 2 wrappers; `plot_stability_heatmap`; `plot_components` (`stability`) |
| Summarization | `summarize_components` (6 `summary_type` values) |
| Jackstraw | `jackstraw_rajive`; `print`, `summary`, `autoplot`, `fortify`; `plot_jackstraw` (3 types); `plot_components` (`volcano`, `jackstraw_summary`); `get_significant_vars`; `summarize_significant_vars` |
| Reporting / export | `export_results`, `rajive_report` |

---

## Prerequisites

```{r}
library(rajiveplus)
library(ggplot2)
```


---

## Shared Simulated Data

```{r}
# ===== Run once (run_heavy = TRUE) to generate the gallery cache =====
set.seed(2)
n   <- 80L
pks <- c(60L, 50L, 40L)

sim      <- ajive.data.sim(K = 3, rankJ = 2, rankA = c(5L, 4L, 4L),
                           n = n, pks = pks, dist.type = 1)
blocks   <- sim$sim_data
names(blocks) <- c("mRNA", "Methylation", "Proteomics")
for (k in seq_along(blocks))
  rownames(blocks[[k]]) <- paste0("sample", seq_len(n))

meta_df <- data.frame(
  group      = rep(c("A", "B"), each = n / 2L),
  score_cont = rnorm(n),
  time_event = abs(rnorm(n, mean = 5, sd = 2)),
  event_flag = rbinom(n, 1L, 0.6),
  row.names  = paste0("sample", seq_len(n)),
  stringsAsFactors = FALSE
)

fit <- Rajive(
  blocks               = blocks,
  initial_signal_ranks = c(5L, 4L, 4L),
  n_wedin_samples      = 200L,
  n_rand_dir_samples   = 200L
)

set.seed(42)
js <- jackstraw_rajive(fit, blocks, n_null = 5L, correction = "BH")

assoc_cat  <- associate_components(fit, meta_df, variable = "group",
                                   mode = "categorical")
assoc_cont <- associate_components(fit, meta_df, variable = "score_cont",
                                   mode = "continuous")

stab_rank <- assess_stability(fit, blocks, c(5L, 4L, 4L),
                              target = "joint_rank", B = 5L)
stab_load <- assess_stability(fit, blocks, c(5L, 4L, 4L),
                              target = "loadings",   B = 5L)
stab_comp <- assess_stability(fit, blocks, c(5L, 4L, 4L),
                              target = "components", B = 5L)

dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
saveRDS(
  list(blocks = blocks, meta = meta_df, fit = fit, js = js,
       assoc_cat = assoc_cat, assoc_cont = assoc_cont,
       stab_rank = stab_rank, stab_load = stab_load, stab_comp = stab_comp),
  file = file.path(data_dir, "gallery_precomp.rds")
)
```

```{r}
# ===== Load cached results (always runs) =====
cache      <- readRDS(file.path(data_dir, "gallery_precomp.rds"))
blocks     <- cache$blocks
meta_df    <- cache$meta
fit        <- cache$fit
js         <- cache$js
assoc_cat  <- cache$assoc_cat
assoc_cont <- cache$assoc_cont
stab_rank  <- cache$stab_rank
stab_load  <- cache$stab_load
stab_comp  <- cache$stab_comp
```


---

# 1  Data Simulation

## `ajive.data.sim()`

Synthetic data give us a model with known structure. That makes it possible to
check whether a function is recovering the planted joint and individual ranks
rather than merely producing plausible-looking output.

```{r}
set.seed(7)
sim2 <- ajive.data.sim(
  K         = 2,
  rankJ     = 2,
  rankA     = c(4, 3),
  n         = 60,
  pks       = c(80, 60),
  dist.type = 1,
  noise     = 1
)
cat("True joint rank:      ", sim2$rankJ_true, "\n")
cat("True individual ranks:", sim2$rankA_true, "\n")
cat("Block 1 dimensions:   ", dim(sim2$sim_data[[1]]), "\n")
```


**What to look for.** `sim_data` is a list of K sample-by-feature matrices.
`rankJ_true` and `rankA_true` record the planted ranks, which become the truth
set for recovery checks.

---

# 2  Core Decomposition

## `Rajive()`

`Rajive()` is the central decomposition. It separates each block into shared
joint structure, block-specific individual structure, and residual noise.

```{r}
fit_demo <- Rajive(
  blocks               = blocks,
  initial_signal_ranks = c(5L, 4L, 4L),
  full                 = TRUE,
  n_wedin_samples      = 200L,
  n_rand_dir_samples   = 200L,
  joint_rank           = NA,
  num_cores            = 1L
)
```


**What to look for.** The result is a `rajive` S3 object. Use
`joint_rank = NA` when the rank should be estimated. Choose
`initial_signal_ranks` conservatively above the expected signal rank in each
block, because this is the search space from which joint structure is found.

---

## `Rajive()` — permutation-based rank selection (`n_perm_samples`)

### What it changes

The default rank-selection pipeline combines two bounds:

1. **Wedin bound** — resamples row-subsets of each block's SVD to estimate
   the perturbation threshold on the joint singular values.
2. **Random-direction bound** — projects random Gaussian vectors through the
   signal subspaces to characterise the null distribution parametrically.

`n_perm_samples` adds a non-parametric alternative to the random-direction
step. In each draw, every block's signal scores are row-shuffled, the blocks
are stacked, and the leading squared singular value is recorded. The 95th
percentile becomes the permutation threshold.

### When to use it

| Scenario | Recommendation |
|---|---|
| Data are approximately Gaussian | Default (`n_rand_dir_samples`) is faster and equally valid |
| Heavy-tailed, count, or compositional data | `n_perm_samples` adapts to the actual noise distribution |
| Strong batch effects within blocks | `n_perm_samples` captures inter-block confounding that Gaussian null may miss |
| Quick exploratory runs | `n_rand_dir_samples` is cheaper per sample |

### Usage

Set `n_perm_samples` to a positive integer and leave `n_rand_dir_samples = NA`
(or any value — permutation takes precedence when both are set).

```{r}
set.seed(42L)
fit_perm <- Rajive(
  blocks               = blocks,
  initial_signal_ranks = c(5L, 4L, 4L),
  full                 = TRUE,
  n_wedin_samples      = 200L,
  n_perm_samples       = 200L,   # replaces random-direction bound
  joint_rank           = NA,
  num_cores            = 1L
)

# Inspect the permutation null
perm_info <- fit_perm$joint_rank_sel$perm
cat("Permutation threshold (95th pctile sq SV):",
    round(perm_info$perm_svsq_threshold, 4), "\n")
cat("Combined threshold (max wedin, perm):      ",
    round(fit_perm$joint_rank_sel$overall_sv_sq_threshold, 4), "\n")
cat("Estimated joint rank:", fit_perm$joint_rank, "\n")
```


**What to look for.** `fit_perm$joint_rank_sel$perm` contains two elements:
`perm_samples` (the vector of `n_perm_samples` squared singular values) and
`perm_svsq_threshold` (the 95th percentile used as the null cut-off). The
combined threshold is `max(wedin_threshold, perm_threshold)` — a conservative
intersection that requires the observed signal to exceed both independent criteria.

### Visualising the permutation null

```{r}
# Diagnostic plot: shows obs sq-SVs against both Wedin and permutation thresholds
plot_components(fit_perm, plot_type = "ajive_diagnostic")
```


The diagnostic panel puts the decision on one page:

- **Observed squared singular values** (dots) with the rank cut-off (vertical
  dashed line).
- **Wedin null distribution** (blue histogram, lower left) — resampled from
  data geometry.
- **Permutation null distribution** (green histogram, lower right) — derived
  by row-shuffling signal scores.

The dashed vertical line in each histogram marks the corresponding 95th-percentile
threshold; the overall cut-off (dashed red line on the SV plot) is their maximum.

### Reproducibility

Permutation draws use R's global RNG. Set `set.seed()` before calling
`Rajive()` when the draws must be reproducible. The samples are also stored in
`fit$joint_rank_sel$perm$perm_samples`, so the null can be inspected later
without refitting.

---

# 3  S3 Methods — `rajive` class

## `print.rajive()`

```{r}
print(fit)
```


**What to look for.** The print method gives the number of blocks, the
estimated joint rank, and the individual rank per block. A joint rank of 0
means the fit did not detect shared signal.

---

## `summary.rajive()`

```{r}
summary(fit)
```


**What to look for.** The summary is a tidy one-row-per-block table. It is the
quick sanity check before extracting scores, loadings, or matrices.

---

# 4  Rank Accessors

## `get_joint_rank()`

```{r}
jrank <- get_joint_rank(fit)
cat("Estimated joint rank:", jrank, "\n")
```


---

## `get_individual_rank()`

```{r}
for (k in seq_along(blocks))
  cat(sprintf("Block %d (%s) individual rank: %d\n",
              k, names(blocks)[k], get_individual_rank(fit, k)))
```


---

## `get_all_ranks()`

```{r}
rank_tbl <- get_all_ranks(fit)
knitr::kable(rank_tbl, caption = "All block ranks")
```


**What to look for.** This combines the shared rank and each block's individual
rank into one reportable table.

---

# 5  Component Extractors

## `get_joint_scores()`

```{r}
J_scores <- get_joint_scores(fit)
cat("Dimensions (samples x joint components):", dim(J_scores), "\n")
print(round(head(J_scores, 4), 3))
```


**What to look for.** Each column is one joint factor and each row is one
sample. These scores are estimated, so downstream association p-values should
be read as exploratory unless uncertainty is propagated.

---

## `get_block_scores()`

```{r}
bl1_joint  <- get_block_scores(fit, k = 1, type = "joint")
bl2_indiv  <- get_block_scores(fit, k = 2, type = "individual")
cat("Block 1 joint scores:", dim(bl1_joint), "\n")
cat("Block 2 individual scores:", dim(bl2_indiv), "\n")
```


---

## `get_block_loadings()`

```{r}
bl1_load <- get_block_loadings(fit, k = 1, type = "joint")
cat("Block 1 joint loadings (features x components):", dim(bl1_load), "\n")
```


**What to look for.** Large absolute loadings mark features aligned with the
component. For feature-level significance, use `jackstraw_rajive()` rather than
the raw loading rank alone.

---

## `get_block_matrix()`

```{r}
J1 <- get_block_matrix(fit, k = 1, type = "joint")
I1 <- get_block_matrix(fit, k = 1, type = "individual")
E1 <- get_block_matrix(fit, k = 1, type = "noise")
cat("Reconstruction error:", round(max(abs(blocks[[1]] - J1 - I1 - E1)), 6), "\n")
```


**What to look for.** The joint, individual, and noise matrices reconstruct the
original block. The split lets you ask whether a pattern is shared across
blocks or belongs mainly to one block.

---

# 6  Diagnostic Extraction

## `extract_components()` — `what = "rank_diagnostics"` (wide)

```{r}
diag_wide <- extract_components(fit, what = "rank_diagnostics", format = "wide")
cat("Cutoff rule:", diag_wide$cutoff_rule, "\n")
cat("Wedin threshold:", round(diag_wide$wedin_cutoff, 4), "\n")
cat("Rand-dir threshold:", round(diag_wide$rand_cutoff, 4), "\n")
cat("Joint rank estimate:", diag_wide$joint_rank_estimate, "\n")
```


**What to look for.** The overall threshold is `max(wedin_cutoff, rand_cutoff)`.
It is a useful empirical rule, not a formal FWER or FDR guarantee.

---

## `extract_components()` — `what = "rank_diagnostics"` (long)

```{r}
diag_long <- extract_components(fit, what = "rank_diagnostics", format = "long")
knitr::kable(head(diag_long, 6),
             caption = "Rank diagnostics — long format (first 6 SVs)")
```


---

## `extract_components()` — `what = "scores"`

```{r}
scores_long <- extract_components(fit, what = "scores", type = "joint",
                                  format = "long")
cat("Columns:", names(scores_long), "\n")
head(scores_long, 4)
```


---

## `extract_components()` — `what = "loadings"`

```{r}
load_long <- extract_components(fit, what = "loadings", type = "joint",
                                block = 1L, format = "long")
cat("Loadings dimensions:", dim(load_long), "\n")
head(load_long, 4)
```


---

## `extract_components()` — `what = "variance"`

```{r}
var_df <- extract_components(fit, blocks = blocks, what = "variance")
knitr::kable(round(var_df, 3), caption = "Variance explained per block (%)")
```


---

# 7  Decomposition Visualization

## `data_heatmap()` — requires *reshape2*

```{r}
if (requireNamespace("reshape2", quietly = TRUE)) {
  data_heatmap(
    data  = t(blocks[[1]][, 1:20]),
    title = "Block 1 (mRNA): first 20 features",
    xlab  = "Samples",
    ylab  = "Features"
  )
}
```


**What to look for.** Coherent color bands across samples point to latent
structure before decomposition.

---

## `decomposition_heatmaps_robustH()` — requires *cowplot* + *reshape2*

```{r}
if (requireNamespace("cowplot", quietly = TRUE) &&
    requireNamespace("reshape2", quietly = TRUE)) {
  decomposition_heatmaps_robustH(blocks = blocks, jive_results_robust = fit)
}
```


**What to look for.** The four panels separate original, joint, individual, and
noise signal. Shared structure should appear in the joint panels across blocks;
the noise panels should be comparatively featureless.

---

# 8  Variance Explained

## `showVarExplained_robust()`

```{r}
var_list <- showVarExplained_robust(fit, blocks)
cat("Joint:     ", round(unlist(var_list$Joint), 3), "\n")
cat("Individual:", round(unlist(var_list$Indiv), 3), "\n")
cat("Residual:  ", round(unlist(var_list$Resid), 3), "\n")
```


---

## `plot_variance_explained()`

```{r}
plot_variance_explained(fit, blocks)
```


**What to look for.** Blocks with a large joint fraction are strongly tied to
the shared signal; blocks with large individual fractions carry more
view-specific biology.

---

# 9  Score Plots

## `plot_scores()`

```{r}
jr <- get_joint_rank(fit)
if (jr >= 2) {
  plot_scores(fit, k = 1, type = "joint", comp_x = 1, comp_y = 2,
              group = meta_df$group)
} else {
  plot_scores(fit, k = 1, type = "joint", comp_x = 1,
              group = meta_df$group)
}
```


**What to look for.** Group separation along a score axis suggests the latent
component tracks that metadata variable.

---

## `plot_components()` — `"pairs"`

```{r}
plot_components(fit, plot_type = "pairs", type = "joint",
                group = meta_df$group)
```


---

## `plot_components()` — `"density"`

```{r}
plot_components(fit, plot_type = "density", type = "joint",
                group = meta_df$group)
```


**What to look for.** Bimodal densities suggest discrete sample structure.
Unimodal densities point more toward a continuous gradient.

---

# 10  AJIVE Diagnostic Plots

## `plot_components()` — `"rank_threshold"`

```{r}
plot_components(fit, plot_type = "rank_threshold") +
  labs(subtitle = "Threshold = max(Wedin, random-direction) — heuristic, not FWER/FDR-controlled")
```


**What to look for.** Squared singular values above the dashed line are called
joint. A visible gap around the cutoff is more reassuring than a borderline
crossing.

---

## `plot_components()` — `"bound_distributions"`

```{r}
plot_components(fit, plot_type = "bound_distributions")
```


**What to look for.** Narrow bound distributions give a stable threshold;
wide or overlapping distributions suggest the rank decision is fragile.

---

## `plot_components()` — `"ajive_diagnostic"` — requires *patchwork*

```{r}
if (requireNamespace("patchwork", quietly = TRUE)) {
  plot_components(fit, plot_type = "ajive_diagnostic")
}
```


**What to look for.** This is the first diagnostic to inspect. Joint components
should clear both cutoffs, and the bound distributions should not make the
decision look accidental.

---

# 11  Feature Visualization

## `plot_components()` — `"top_features"`

```{r}
plot_components(fit, plot_type = "top_features", type = "joint",
                block = 1L, component = 1L, top_n = 20L)
```


**What to look for.** The top features drive the plotted component most
strongly. Treat this as ranking, not as significance.

---

## `plot_components()` — `"component_heatmap"`

```{r}
plot_components(fit, plot_type = "component_heatmap", type = "joint",
                block = 1L)
```


**What to look for.** Groups of features with similar loading patterns suggest
modules that move together across the component.

---

# 12  S3 autoplot / fortify — `rajive` class

## `autoplot()`

```{r}
autoplot(fit)
```


---

## `fortify()`

```{r}
tidy_fit <- fortify(fit)
cat("Columns:", names(tidy_fit), "\n")
head(tidy_fit, 3)
```


**What to look for.** `fortify()` returns one row per sample per joint
component, ready for `ggplot2` or `dplyr`.

---

# 13  Feature Ranking Suite

## `rank_features()` — `"top_loadings"`

```{r}
top20 <- rank_features(fit, mode = "top_loadings", type = "joint",
                       block = 1L, component = 1L, n = 20L, signed = TRUE)
knitr::kable(head(top20, 10), digits = 4,
             caption = "Top 10 features — Block 1, Joint Component 1")
```


**What to look for.** High absolute loadings are the strongest component
drivers in this block. Use `jackstraw_rajive()` when the question is
statistical significance.

---

## `rank_features()` — `"contribution"`

```{r}
contrib <- rank_features(fit, mode = "contribution", type = "joint",
                         block = 1L, component = 1L)
knitr::kable(head(contrib, 8), digits = 4,
             caption = "Feature contributions (top 8)")
```


---

## `rank_features()` — `"overlap"`

```{r}
overlap <- rank_features(fit, mode = "overlap", type = "joint",
                         component = 1L, n = 50L, method = "jaccard")
knitr::kable(overlap, digits = 3,
             caption = "Jaccard overlap of top-50 joint features")
```


**What to look for.** Values near 1 mean the same features dominate both
blocks. Moderate overlap is common when different molecular layers share a
process without sharing identical feature identities.

---

### Wrappers: `get_top_loadings()`, `get_feature_contributions()`, `compare_feature_sets_across_blocks()`

```{r}
top_w     <- get_top_loadings(fit, block = 1L, component = 1L, n = 5L)
overlap_w <- compare_feature_sets_across_blocks(fit, component = 1L,
                                                n = 50L, method = "jaccard")
cat("get_top_loadings — first row:\n"); print(top_w[1, ])
cat("compare_feature_sets_across_blocks:\n"); print(overlap_w)
```


---

# 14  Association Testing Suite

## `associate_components()` — `"categorical"`

```{r}
knitr::kable(assoc_cat, digits = 4,
             caption = "Joint factor — group association (Kruskal-Wallis)")
```


> **Exploratory-inference caveat.** These p-values condition on estimated
> scores and do not propagate score uncertainty. Treat them as
> hypothesis-generating.

---

## `associate_components()` — `"continuous"`

```{r}
knitr::kable(assoc_cont, digits = 4,
             caption = "Joint factor — continuous association (Pearson)")
```


---

## `associate_components()` — `"survival"` (requires *survival*)

```{r}
assoc_surv <- associate_components(
  fit, meta_df,
  mode       = "survival",
  time_col   = "time_event",
  status_col = "event_flag",
  split      = "median"
)
knitr::kable(assoc_surv, digits = 4,
             caption = "Joint factor — survival association (log-rank, median split)")
```


> **Survival-split note.** The median-split log-rank test is descriptive. For
> primary inference, prefer `split = "none"` and the continuous Cox model.

---

## `plot_components()` — `"association"`

```{r}
assoc_all <- rbind(assoc_cat, assoc_cont)
plot_components(assoc_results = assoc_all, plot_type = "association")
```


---

### Wrappers: `associate_scores_continuous()`, `associate_scores_categorical()`

```{r}
cont_w <- associate_scores_continuous(fit, meta_df, variable = "score_cont")
cat("Same as associate_components(mode='continuous'):\n")
print(cont_w[1:2, ])
```


---

# 15  Stability Assessment Suite

## `assess_stability()` — `"joint_rank"`

```{r}
cat("Bootstrap joint rank distribution (B=5):\n")
print(stab_rank$rank_table)
cat("Observed rank:", stab_rank$observed_rank, "\n")
```


**What to look for.** A concentrated bootstrap distribution means the selected
rank is insensitive to sample resampling.

---

## `assess_stability()` — `"loadings"`

```{r}
for (k in seq_along(stab_load))
  cat(sprintf("  %s: %s\n", names(stab_load)[k],
              paste(round(stab_load[[k]]$cos_similarity, 3), collapse = ", ")))
```


> **Procrustes note.** Bootstrap loadings are aligned by orthogonal Procrustes
> rotation before cosine similarities are computed. Cosine similarity above
> 0.9 is typically a stable direction.

---

## `assess_stability()` — `"components"`

```{r}
knitr::kable(stab_comp, digits = 3,
             caption = "Bootstrap stability of joint scores (B=5)")
```


---

## `plot_stability_heatmap()` / `plot_components(plot_type = "stability")`

```{r}
plot_stability_heatmap(stab_load)
```

```{r}
plot_components(stability_result = stab_load, plot_type = "stability")
```


---

### Wrappers: `bootstrap_joint_rank()`, `bootstrap_loading_stability()`

```{r}
rank_draws <- bootstrap_joint_rank(blocks, c(5L, 4L, 4L), B = 5L)
load_stab  <- bootstrap_loading_stability(fit, blocks, c(5L, 4L, 4L), B = 5L)
```


---

# 16  Summarization Suite

## `summarize_components()` — `"components"`

```{r}
knitr::kable(summarize_components(fit, summary_type = "components"),
             caption = "Component catalog")
```


---

## `summarize_components()` — `"variance"`

```{r}
knitr::kable(
  summarize_components(fit, blocks = blocks, summary_type = "variance"),
  digits = 3,
  caption = "Variance decomposition per block"
)
```


---

## `summarize_components()` — `"associations"`

```{r}
knitr::kable(
  summarize_components(assoc_results = assoc_all, summary_type = "associations"),
  digits = 4,
  caption = "Association summary"
)
```


---

## `summarize_components()` — `"diagnostics"`

```{r}
knitr::kable(head(summarize_components(fit, summary_type = "diagnostics"), 6),
             digits = 4, caption = "Rank diagnostics (first 6 rows)")
```


---

## `summarize_components()` — `"stability"`

```{r}
knitr::kable(
  summarize_components(stability_result = stab_rank, summary_type = "stability"),
  digits = 3, caption = "Bootstrap joint rank distribution"
)
```


---

## `summarize_components()` — `"significance"`

```{r}
knitr::kable(
  summarize_components(jackstraw_result = js, summary_type = "significance"),
  digits = 4, caption = "Significant features per block x component"
)
```


---

# 17  Jackstraw Significance Testing

## `jackstraw_rajive()`

```{r}
set.seed(42)
js <- jackstraw_rajive(
  ajive_output = fit,
  blocks       = blocks,
  alpha        = 0.05,
  n_null       = 5L,       # use >= 100 in practice
  correction   = "BH"
)
```


> **Significance note.** The jackstraw identifies features associated with the
> estimated joint score. It does not establish causal drivers, and the
> decomposition is not re-estimated after feature masking.

---

## `print.jackstraw_rajive()` / `summary.jackstraw_rajive()`

```{r}
print(js)
summary(js)
```


---

## `autoplot.jackstraw_rajive()` / `fortify.jackstraw_rajive()`

```{r}
autoplot(js)
```

```{r}
tidy_js <- fortify(js)
cat("Columns:", names(tidy_js), "\n")
head(tidy_js, 4)
```


---

## `plot_jackstraw()` — `"pvalue_hist"`

```{r}
plot_jackstraw(js, type = "pvalue_hist", block = 1L, component = 1L)
```


**What to look for.** A spike near zero followed by a near-uniform tail suggests
real signal plus calibrated null features. A fully uniform histogram suggests
little feature-level signal for that component.

---

## `plot_jackstraw()` — `"scatter"`

```{r}
plot_jackstraw(js, type = "scatter", block = 1L, component = 1L,
               label_top = 5L)
```


---

## `plot_jackstraw()` — `"loadings_significance"`

```{r}
plot_jackstraw(js, type = "loadings_significance", block = 1L)
```


---

## `plot_components()` — `"volcano"` / `"jackstraw_summary"`

```{r}
plot_components(jackstraw_result = js, plot_type = "volcano",
                block = 1L, component = 1L)
```

```{r}
plot_components(jackstraw_result = js, plot_type = "jackstraw_summary")
```


---

## `get_significant_vars()` / `summarize_significant_vars()`

```{r}
sig_b1c1 <- get_significant_vars(js, block = 1L, component = 1L)
cat("Significant features (block 1, component 1):", nrow(sig_b1c1), "\n")
if (nrow(sig_b1c1) > 0)
  knitr::kable(head(sig_b1c1, 8), digits = 4)
```

```{r}
sig_all <- summarize_significant_vars(js)
knitr::kable(head(sig_all, 8), digits = 4,
             caption = "All significant features (all blocks x components)")
```


---

# 18  Reporting and Export

## `export_results()` — `"csv"` / `"rds"`

```{r}
top_feats <- rank_features(fit, mode = "top_loadings", type = "joint",
                           block = 1L, component = 1L, n = 20L)
tmp_csv <- tempfile(fileext = ".csv")
export_results(top_feats, path = tmp_csv, format = "csv")
rt <- read.csv(tmp_csv)
cat("Round-trip rows/cols:", nrow(rt), "/", ncol(rt), "\n")
```

```{r}
tmp_rds <- tempfile(fileext = ".rds")
export_results(fit, path = tmp_rds, format = "rds")
fit2 <- readRDS(tmp_rds)
cat("Reloaded joint rank:", get_joint_rank(fit2), "\n")
```


---

## `rajive_report()`

```{r}
tmp_html <- tempfile(fileext = ".html")
rajive_report(
  ajive_output     = fit,
  blocks           = blocks,
  metadata         = meta_df,
  jackstraw_result = js,
  output_file      = tmp_html,
  sections         = c("overview", "variance", "features",
                       "associations", "stability")
)
cat("Report written to:", tmp_html, "\n")
cat("File size (bytes):", file.info(tmp_html)$size, "\n")
```


**What to look for.** The report is a portable review artifact. For final
analysis decisions, use the specialized tables and plots shown above.

---

# 19  Appendix — Wrapper Quick Reference

| Wrapper | Equivalent underlying call |
|---|---|
| `get_top_loadings(fit, block=1, component=1)` | `rank_features(fit, mode="top_loadings", block=1, component=1)` |
| `get_feature_contributions(fit, block=1)` | `rank_features(fit, mode="contribution", block=1)` |
| `compare_feature_sets_across_blocks(fit, n=50)` | `rank_features(fit, mode="overlap", n=50)` |
| `summarize_significant_vars(js)` | `rank_features(jackstraw_result=js, mode="significant")` |
| `associate_scores_continuous(fit, meta, variable="x")` | `associate_components(fit, meta, variable="x", mode="continuous")` |
| `associate_scores_categorical(fit, meta, variable="g")` | `associate_components(fit, meta, variable="g", mode="categorical")` |
| `associate_scores_survival(fit, meta, time_col="t", status_col="e")` | `associate_components(fit, meta, mode="survival", time_col="t", status_col="e")` |
| `bootstrap_joint_rank(blocks, ranks, B=5)` | `assess_stability(blocks=blocks, initial_signal_ranks=ranks, target="joint_rank", B=5)` |
| `bootstrap_loading_stability(fit, blocks, ranks, B=5)` | `assess_stability(fit, blocks, ranks, target="loadings", B=5)` |
| `plot_stability_heatmap(stab_load)` | `plot_components(stability_result=stab_load, plot_type="stability")` |

---

## Session Info

```{r}
sessionInfo()
```
