---
title: "Inference for Joint Structure in Simulated RaJIVE Data"
author: "rajiveplus package authors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Inference for Joint Structure in Simulated RaJIVE Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

The decomposition of multiview data into shared and view-specific structure is
useful only to the extent that the resulting components can be interpreted. A
joint component may be visually convincing, yet its rank, its feature loadings
and its association with an external variable remain estimates. In practice,
these estimates vary with the samples that happen to be observed.

Here we use a small simulated data set to show how `rajiveplus` places
uncertainty around the main quantities reported by `Rajive()`. The bootstrap
counts are kept deliberately small so that the vignette can be checked with the
package. Larger values of `B` should be used when the intervals themselves are
part of a scientific analysis.

```{r setup}
library(rajiveplus)
```

## Simulate and fit

```{r simulate-fit}
set.seed(101)
n <- 64L
pks <- c(10L, 8L)
sim <- ajive.data.sim(K = 2, rankJ = 1L, rankA = c(1L, 1L),
                      n = n, pks = pks, dist.type = 1)
blocks <- sim$sim_data
names(blocks) <- c("rna", "protein")
for (k in seq_along(blocks)) {
  rownames(blocks[[k]]) <- paste0("sample", seq_len(n))
  colnames(blocks[[k]]) <- paste0(names(blocks)[[k]], "_", seq_len(ncol(blocks[[k]])))
}

fit <- Rajive(blocks, initial_signal_ranks = c(2L, 2L),
              joint_rank = 1L,
              n_wedin_samples = NA,
              n_rand_dir_samples = NA)

stopifnot(inherits(fit, "rajive"))
stopifnot(fit$joint_rank == 1L)
```

## Feature-level variance partition

One interpretation of a RaJIVE fit is that each feature contains a mixture of
shared signal, block-specific signal and residual variation. The function
`joint_variance_partition()` makes this interpretation explicit by assigning
each feature's sum of squares to the joint, individual and residual terms. The
residual term is computed after reconstructing the fitted joint and individual
matrices, and therefore the original data blocks are supplied along with the
fit.

```{r variance-partition}
vp <- joint_variance_partition(fit, blocks)
head(vp)

stopifnot(all(c("joint_prop", "individual_prop", "residual_prop") %in% names(vp)))
stopifnot(all(is.finite(vp$joint_prop)))
```

## Loading confidence intervals

Feature loadings are often the first route from a latent component back to
biology. The difficulty is that a loading is not fixed by the algorithm alone:
it also depends on the sampled cells or specimens. The bootstrap samples below
refit the model and align each result to the loading matrix estimated from the
full data before intervals are formed. The intervals are entry-wise summaries,
so nearly tied components should still be interpreted at the level of directions
or subspaces rather than as isolated coefficients.

```{r loading-ci}
set.seed(102)
ci_load <- rajive_ci(fit, blocks, c(2L, 2L),
                     target = "loadings",
                     method = "percentile",
                     B = 20L,
                     joint_rank = 1L,
                     n_wedin_samples = NA,
                     n_rand_dir_samples = NA)

load_tbl <- extract_components(fit, what = "loadings",
                               format = "long", ci = ci_load)
head(load_tbl)

stopifnot(all(c("lower", "upper", "ci_method") %in% names(load_tbl)))
```

```{r loading-ci-plot}
plot_components(fit, plot_type = "loading_ci", ci = ci_load, top_n = 25L)
```

## BCa rank interval

The selected joint rank is also an estimate. A percentile interval describes
the empirical bootstrap distribution, while a BCa interval additionally uses a
jackknife estimate of acceleration. For observation-level data this means
leaving out one sample at a time; when a cluster label is supplied, the same
principle is applied to clusters. The price is computational, because the model
must be refit for each jackknife deletion.

```{r rank-bca}
set.seed(103)
ci_rank <- rajive_ci(fit, blocks, c(2L, 2L),
                     target = "joint_rank",
                     method = "bca",
                     B = 20L,
                     joint_rank = 1L,
                     n_wedin_samples = NA,
                     n_rand_dir_samples = NA)
ci_rank

stopifnot(ci_rank$method == "bca")
stopifnot(is.finite(ci_rank$lower), is.finite(ci_rank$upper))
```

## Score-uncertainty propagation

Associations between component scores and metadata are commonly used to attach
phenotype to a latent axis. Such an association is conditional on the estimated
scores. With `propagate_uncertainty = "bootstrap"`, `associate_components()`
keeps the usual point-estimate p value and adds three summaries from bootstrap
refits: the fraction of refits that remain significant, an interval for the
effect size, the median bootstrap p value, and the number of valid bootstrap
refits contributing to each row.

```{r association-propagation}
set.seed(104)
metadata <- data.frame(marker = fit$joint_scores[, 1L] + rnorm(n, sd = 0.05))

assoc <- associate_components(
  fit, metadata,
  variable = "marker",
  mode = "continuous",
  propagate_uncertainty = "bootstrap",
  blocks = blocks,
  initial_signal_ranks = c(2L, 2L),
  B = 20L,
  joint_rank = 2L,
  n_wedin_samples = NA,
  n_rand_dir_samples = NA
)
assoc

stopifnot(all(c("stability", "effect_lo", "effect_hi", "p_median",
                "n_boot_valid") %in% names(assoc)))
```

For screening across the whole decomposition, `associate_all_components()`
keeps joint and block-specific individual score families explicit. Bootstrap
replicates for each family are aligned to that family's full-data fitted basis;
individual component summaries should therefore be read as basis-conditional
sensitivity summaries rather than formal post-selection inference.

```{r association-all-components}
set.seed(105)
metadata_i <- data.frame(marker = fit$block_decomps[[1L]]$u[, 1L] +
                           rnorm(n, sd = 0.05))

assoc_all <- associate_all_components(
  fit, metadata_i,
  variable = "marker",
  mode = "continuous",
  include = "both",
  blocks_to_include = 1L,
  propagate_uncertainty = "bootstrap",
  blocks = blocks,
  initial_signal_ranks = c(2L, 2L),
  B = 6L,
  joint_rank = 2L,
  n_wedin_samples = NA,
  n_rand_dir_samples = NA
)
assoc_all

stopifnot(all(c("source", "block", "stability", "n_boot_valid") %in%
                names(assoc_all)))
```

The association table can be visualized either as an adjusted-p-value heatmap
or, when bootstrap propagation is available, as an uncertainty forest showing
the point-estimate effect, bootstrap effect interval, stability, and number of
valid refits.

```{r association-visualization, fig.width=7, fig.height=4}
plot_components(assoc_results = assoc_all, plot_type = "association")

plot_components(assoc_results = assoc_all, plot_type = "association",
                style = "uncertainty")
```
