---
title: "Bootstrap Inference on Simulated RaJIVE Data"
author: "rajiveplus package authors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Bootstrap Inference on Simulated RaJIVE Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

This vignette is a compact end-to-end check of the inference helpers added on
top of `Rajive()`. It uses simulated data and intentionally small bootstrap
counts so it can run during package tests. Increase `B` for real analyses.

```{r setup}
library(rajiveplus)
```

## Simulate and Fit

```{r simulate-fit}
set.seed(101)
n <- 64L
pks <- c(10L, 8L)
sim <- ajive.data.sim(K = 2, rankJ = 1L, rankA = c(1L, 1L),
                      n = n, pks = pks, dist.type = 1)
blocks <- sim$sim_data
names(blocks) <- c("rna", "protein")
for (k in seq_along(blocks)) {
  rownames(blocks[[k]]) <- paste0("sample", seq_len(n))
  colnames(blocks[[k]]) <- paste0(names(blocks)[[k]], "_", seq_len(ncol(blocks[[k]])))
}

fit <- Rajive(blocks, initial_signal_ranks = c(2L, 2L),
              joint_rank = 1L,
              n_wedin_samples = NA,
              n_rand_dir_samples = NA)

stopifnot(inherits(fit, "rajive"))
stopifnot(fit$joint_rank == 1L)
```

## Feature-Level Variance Partition

`joint_variance_partition()` reports how much of each feature's sum of squares
is attributed to joint, individual, and residual structure. The residual is
computed from the original data minus reconstructed joint and individual
matrices, so the original `blocks` are required.

```{r variance-partition}
vp <- joint_variance_partition(fit, blocks)
head(vp)

stopifnot(all(c("joint_prop", "individual_prop", "residual_prop") %in% names(vp)))
stopifnot(all(is.finite(vp$joint_prop)))
```

## Loading Confidence Intervals

Bootstrap loading intervals are aligned to the full-data loading matrix before
summarisation. These are entry-wise intervals; loading directions and subspaces
remain the statistically safer interpretation when components are near-tied.

```{r loading-ci}
set.seed(102)
ci_load <- rajive_ci(fit, blocks, c(2L, 2L),
                     target = "loadings",
                     method = "percentile",
                     B = 20L,
                     joint_rank = 1L,
                     n_wedin_samples = NA,
                     n_rand_dir_samples = NA)

load_tbl <- extract_components(fit, what = "loadings",
                               format = "long", ci = ci_load)
head(load_tbl)

stopifnot(all(c("lower", "upper", "ci_method") %in% names(load_tbl)))
```

```{r loading-ci-plot}
plot_components(fit, plot_type = "loading_ci", ci = ci_load, top_n = 25L)
```

## BCa Rank Interval

BCa intervals use delete-one observation jackknife acceleration, or delete-one
cluster acceleration when `cluster` is supplied. This is more expensive than
percentile intervals because it refits `Rajive()` for every jackknife deletion.

```{r rank-bca}
set.seed(103)
ci_rank <- rajive_ci(fit, blocks, c(2L, 2L),
                     target = "joint_rank",
                     method = "bca",
                     B = 20L,
                     joint_rank = 1L,
                     n_wedin_samples = NA,
                     n_rand_dir_samples = NA)
ci_rank

stopifnot(ci_rank$method == "bca")
stopifnot(is.finite(ci_rank$lower), is.finite(ci_rank$upper))
```

## Score-Uncertainty Propagation

`associate_components()` still returns the point-estimate association p-value.
With `propagate_uncertainty = "bootstrap"`, it also reports the fraction of
bootstrap score refits that remain significant, a bootstrap effect-size interval,
and the median bootstrap p-value.

```{r association-propagation}
set.seed(104)
metadata <- data.frame(marker = fit$joint_scores[, 1L] + rnorm(n, sd = 0.05))

assoc <- associate_components(
  fit, metadata,
  variable = "marker",
  mode = "continuous",
  propagate_uncertainty = "bootstrap",
  blocks = blocks,
  initial_signal_ranks = c(2L, 2L),
  B = 20L,
  joint_rank = 2L,
  n_wedin_samples = NA,
  n_rand_dir_samples = NA
)
assoc

stopifnot(all(c("stability", "effect_lo", "effect_hi", "p_median") %in% names(assoc)))
```
