---
title: "How Certain Is a Joint Component?"
author: "rajiveplus package authors"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{How Certain Is a Joint Component?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

RaJIVE gives a joint component, but the scientific question begins one step
later: how much of that component would survive a new draw of samples? The
rank, feature loadings, variance attribution, and phenotype association are all
estimates. They can look decisive while still being sample-sensitive.

This vignette uses a small simulated data set to put uncertainty around the
main quantities returned by `Rajive()`. The bootstrap counts are intentionally
small so the vignette can be checked with the package. Use larger `B` values
when intervals are part of a real analysis.

```{r setup}
library(rajiveplus)
```

## A small joint signal

```{r simulate-fit}
set.seed(101)
n <- 64L
pks <- c(10L, 8L)
sim <- ajive.data.sim(K = 2, rankJ = 1L, rankA = c(1L, 1L),
                      n = n, pks = pks, dist.type = 1)
blocks <- sim$sim_data
names(blocks) <- c("rna", "protein")
for (k in seq_along(blocks)) {
  rownames(blocks[[k]]) <- paste0("sample", seq_len(n))
  colnames(blocks[[k]]) <- paste0(names(blocks)[[k]], "_", seq_len(ncol(blocks[[k]])))
}

fit <- Rajive(blocks, initial_signal_ranks = c(2L, 2L),
              joint_rank = 1L,
              n_wedin_samples = NA,
              n_rand_dir_samples = NA)

stopifnot(inherits(fit, "rajive"))
stopifnot(fit$joint_rank == 1L)
```

## Where does each feature's variance go?

Each feature can carry shared signal, block-specific signal, and residual
variation at the same time. `joint_variance_partition()` makes that accounting
explicit by assigning the feature-level sum of squares to joint, individual,
and residual terms. The residual is computed after reconstructing the fitted
joint and individual matrices, so the original data blocks are supplied with
the fit.

```{r variance-partition}
vp <- joint_variance_partition(fit, blocks)
head(vp)

stopifnot(all(c("joint_prop", "individual_prop", "residual_prop") %in% names(vp)))
stopifnot(all(is.finite(vp$joint_prop)))
```

## Which loadings are stable?

Loadings are the bridge from a latent component back to biology, but they are
not fixed constants. They depend on the samples used to fit the model. The
bootstrap below refits RaJIVE, aligns each refit to the full-data loading
matrix, and then forms entry-wise intervals. When components are nearly tied,
read these intervals as summaries of directions or subspaces, not as isolated
coefficient certainties.

```{r loading-ci}
set.seed(102)
ci_load <- rajive_ci(fit, blocks, c(2L, 2L),
                     target = "loadings",
                     method = "percentile",
                     B = 20L,
                     joint_rank = 1L,
                     n_wedin_samples = NA,
                     n_rand_dir_samples = NA)

load_tbl <- extract_components(fit, what = "loadings",
                               format = "long", ci = ci_load)
head(load_tbl)

stopifnot(all(c("lower", "upper", "ci_method") %in% names(load_tbl)))
```

```{r loading-ci-plot}
p_load <- plot_components(fit, plot_type = "loading_ci", ci = ci_load,
                          top_n = 25L)
print(p_load)
```

## How stable is the rank?

The joint rank is also estimated. A percentile interval reports the bootstrap
distribution directly. A BCa interval adds a jackknife estimate of acceleration:
one sample is left out at a time, or one cluster at a time when clusters are
supplied. The extra information costs extra refits.

```{r rank-bca}
set.seed(103)
ci_rank <- rajive_ci(fit, blocks, c(2L, 2L),
                     target = "joint_rank",
                     method = "bca",
                     B = 20L,
                     joint_rank = 1L,
                     n_wedin_samples = NA,
                     n_rand_dir_samples = NA)
ci_rank

stopifnot(ci_rank$method == "bca")
stopifnot(is.finite(ci_rank$lower), is.finite(ci_rank$upper))
```

## Do associations survive refitting?

Associations between component scores and metadata attach phenotype to a latent
axis, but they condition on estimated scores. With
`propagate_uncertainty = "bootstrap"`, `associate_components()` keeps the usual
point-estimate p value and adds refit-based summaries: the fraction of
significant refits, an effect-size interval, the median bootstrap p value, and
the number of valid refits behind the row.

```{r association-propagation}
set.seed(104)
metadata <- data.frame(marker = fit$joint_scores[, 1L] + rnorm(n, sd = 0.05))

assoc <- associate_components(
  fit, metadata,
  variable = "marker",
  mode = "continuous",
  propagate_uncertainty = "bootstrap",
  blocks = blocks,
  initial_signal_ranks = c(2L, 2L),
  B = 20L,
  joint_rank = 2L,
  n_wedin_samples = NA,
  n_rand_dir_samples = NA
)
assoc

stopifnot(all(c("stability", "effect_lo", "effect_hi", "p_median",
                "n_boot_valid") %in% names(assoc)))
```

To screen the whole decomposition, `associate_all_components()` keeps joint and
block-specific individual score families separate. Bootstrap replicates are
aligned to each family's full-data basis, so individual-component summaries are
basis-conditional sensitivity checks rather than formal post-selection
inference.

```{r association-all-components}
set.seed(105)
metadata_i <- data.frame(marker = fit$block_decomps[[1L]]$u[, 1L] +
                           rnorm(n, sd = 0.05))

assoc_all <- associate_all_components(
  fit, metadata_i,
  variable = "marker",
  mode = "continuous",
  include = "both",
  blocks_to_include = 1L,
  propagate_uncertainty = "bootstrap",
  blocks = blocks,
  initial_signal_ranks = c(2L, 2L),
  B = 6L,
  joint_rank = 2L,
  n_wedin_samples = NA,
  n_rand_dir_samples = NA
)
assoc_all

stopifnot(all(c("source", "block", "stability", "n_boot_valid") %in%
                names(assoc_all)))
```

The association table has two useful views: an adjusted-p-value heatmap for the
scan, and an uncertainty forest for the point effect, bootstrap interval,
stability, and valid-refit count.

```{r association-visualization, fig.width=7, fig.height=4}
p_assoc <- plot_components(assoc_results = assoc_all,
                           plot_type = "association")
print(p_assoc)

p_unc <- plot_components(assoc_results = assoc_all, plot_type = "association",
                         style = "uncertainty")
print(p_unc)
```
