---
title: "What Do Microbial Kingdoms Share?"
author: "RaJIVEutils development team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{What Do Microbial Kingdoms Share?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  cache      = FALSE
)
data_dir <- "data"
```

# The question

A gut microbiome is not one ecosystem measured three times. Bacteria, fungi,
and viruses can move together, or they can carry kingdom-specific variation.
This vignette uses **rajiveplus** on the multi-kingdom gut microbiome data from
[Haak *et al.* (2021)](https://doi.org/10.1038/s41586-021-03406-7) to separate
those two possibilities.

The cohort contains 57 patients with matched measurements across three
microbial kingdoms - **Bacteria** (16S rRNA, 180 OTUs), **Fungi** (ITS2, 18
OTUs), and **Viruses** (virome sequencing, 42 viral families) - plus clinical
metadata including antibiotic exposure and short-chain fatty acid profiles.

| Kingdom  | Technology    | Features |
|----------|--------------|----------|
| Bacteria | 16S rRNA     | 180 OTUs |
| Fungi    | ITS2         | 18 OTUs  |
| Viruses  | virome-seq   | 42 vOTUs |

## Scientific question

Can RaJIVE find a shared microbiome axis across kingdoms, and can it separate
that axis from bacteria-, fungi-, or virus-specific structure?

# Prerequisites

```{r libs, message=FALSE}
library(rajiveplus)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

theme_set(theme_bw(base_size = 13))
```

# Data loading

The rendered vignette reads cached RDS files. That keeps the article
reproducible without requiring a fresh download from the source archive.

```{r load-raw}
raw  <- readRDS(file.path(data_dir, "microbiome_raw.rds"))
meta <- raw$metadata
dt   <- raw$dt   # long-format: columns sample, feature, view, value

cat("Long-format data :", nrow(dt), "rows x", ncol(dt), "cols\n")
cat("Kingdoms (views) :", paste(sort(unique(dt$view)), collapse = ", "), "\n")
cat("Total samples    :", length(unique(dt$sample)), "\n")
cat("Metadata samples :", nrow(meta), "\n")

# Summarise per-kingdom feature counts
kingdom_summary <- dt |>
  dplyr::group_by(view) |>
  dplyr::summarise(n_samples  = dplyr::n_distinct(sample),
                   n_features = dplyr::n_distinct(feature),
                   .groups    = "drop")
knitr::kable(kingdom_summary, caption = "Features and samples per kingdom")
```

## What scale are the values on?

```{r distributions, fig.height=4}
ggplot(dt, aes(x = value, colour = view, fill = view)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  facet_wrap(~view, scales = "free_y", nrow = 1) +
  labs(title = "CLR-normalised value distributions per kingdom",
       x = "CLR value", y = "Density") +
  theme(legend.position = "none")
```

## Which sample groups are present?

```{r sample-counts}
if ("Category" %in% colnames(meta)) {
  cnt <- sort(table(meta$Category), decreasing = TRUE)
  knitr::kable(as.data.frame(cnt), col.names = c("Category", "Count"),
               caption = "Sample counts per category")
}
```

# Matched blocks

```{r load-preprocessed}
pp      <- readRDS(file.path(data_dir, "microbiome_preprocessed.rds"))
blocks  <- pp$blocks    # named list: Bacteria / Fungi / Viruses
meta_s  <- pp$meta      # metadata aligned to the 57 common samples

cat("Block names :", paste(names(blocks), collapse = ", "), "\n")
cat("Block dims  :",
    paste(sapply(blocks, function(m) paste(dim(m), collapse = "x")), collapse = ", "), "\n")
cat("Meta samples:", nrow(meta_s), "\n")
```

## How much rank should each kingdom offer?

```{r scree, fig.height=4}
svd_list <- readRDS(file.path(data_dir, "microbiome_svd_list.rds"))

scree_df <- do.call(rbind, lapply(names(svd_list), function(k) {
  sv   <- svd_list[[k]]
  rank <- min(length(sv), 15L)
  data.frame(kingdom = k,
             component = seq_len(rank),
             variance  = sv[seq_len(rank)]^2 / sum(sv^2),
             stringsAsFactors = FALSE)
}))

ggplot(scree_df, aes(x = component, y = variance * 100, colour = kingdom)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~kingdom, scales = "free_y") +
  labs(title = "Scree plots — per-kingdom variance explained",
       x = "Principal component", y = "Variance explained (%)") +
  theme(legend.position = "none")
```

# Fit the RaJIVE model

```{r load-fit}
mb_fit   <- readRDS(file.path(data_dir, "microbiome_rajive_results.rds"))
jr       <- get_joint_rank(mb_fit)
all_rnks <- get_all_ranks(mb_fit)   # named by block index "block1","block2","block3"

kingdom_names <- names(blocks)  # "Bacteria","Fungi","Viruses" (from pp$blocks)

cat("Joint rank      :", jr, "\n")
cat("Ranks per block :\n")
print(all_rnks)
```

# Was the joint rank convincing?

```{r rank-diag, fig.height=5}
plot_components(mb_fit, plot_type = "ajive_diagnostic") +
  labs(title = "RaJIVE rank diagnostic")
```

```{r ind-ranks}
ind_rank_df <- data.frame(
  Kingdom       = kingdom_names,
  `Joint rank`  = jr,
  `Indiv. rank` = sapply(seq_along(kingdom_names), function(i)
                    get_individual_rank(mb_fit, i)),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
knitr::kable(ind_rank_df, caption = "Estimated ranks per kingdom")
```

# Where does the variation live?

```{r var-explained, fig.height=4}
plot_variance_explained(mb_fit, blocks) +
  labs(title = "Variance explained — joint vs individual")
```

```{r var-table}
ve_tbl <- showVarExplained_robust(mb_fit, blocks)
if (is.data.frame(ve_tbl)) knitr::kable(ve_tbl, digits = 3,
  caption = "Proportion of variance explained")
```

# What does the joint score track?

```{r joint-scores, fig.height=5}
J_scores <- get_joint_scores(mb_fit)   # n_samples × jr matrix

if (!is.null(J_scores) && jr >= 1) {
  score_df <- as.data.frame(J_scores)
  colnames(score_df) <- paste0("JS", seq_len(jr))
  # J_scores may lack rownames; use meta_s sample identifiers (same ordering)
  score_df$sample <- meta_s$sample

  # Merge with metadata
  meta_sub <- meta_s[, intersect(colnames(meta_s),
                                  c("sample", "Category", "Diagnosis", "Sexs")),
                     drop = FALSE]
  score_df <- merge(score_df, meta_sub, by = "sample", all.x = TRUE)

  if ("Category" %in% colnames(score_df)) {
    ggplot(score_df, aes(x = Category, y = JS1, fill = Category)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 21) +
      geom_jitter(width = 0.15, size = 1.5, alpha = 0.6) +
      labs(title = "Joint score 1 by sample category",
           x = "Category", y = "Joint score 1") +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 30, hjust = 1))
  }
} else {
  message("Joint rank is 0 — no joint scores to plot.")
}
```

## A two-dimensional joint map, if present

```{r joint-scatter, fig.height=5}
if (!is.null(J_scores) && jr >= 2) {
  ggplot(score_df, aes(x = JS1, y = JS2,
                        colour = if ("Category" %in% colnames(score_df))
                                    Category else "grey")) +
    geom_point(size = 3, alpha = 0.8) +
    labs(title = "Joint scores 1 vs 2", colour = "Category")
}
```

# Which features define the joint component?

```{r loadings, fig.height=5}
if (jr >= 1) {
  loading_plots <- lapply(seq_along(kingdom_names), function(i) {
    k  <- kingdom_names[i]
    ld <- tryCatch(
      rank_features(mb_fit, block = i, component = 1L, type = "joint"),
      error = function(e) NULL
    )
    if (is.null(ld)) return(NULL)
    top <- head(ld[order(abs(ld$loading), decreasing = TRUE), ], 15)
    ggplot(top, aes(x = reorder(feature_name, loading), y = loading,
                    fill = loading > 0)) +
      geom_col(width = 0.7) +
      coord_flip() +
      scale_fill_manual(values = c("steelblue", "tomato"), guide = "none") +
      labs(title = paste("Top loadings —", k, "(joint)"),
           x = NULL, y = "Loading")
  })
  loading_plots <- Filter(Negate(is.null), loading_plots)
  for (p in loading_plots) print(p)
}
```

# What is kingdom-specific?

```{r ind-scores, fig.height=5}
for (i in seq_along(kingdom_names)) {
  k  <- kingdom_names[i]
  ir <- get_individual_rank(mb_fit, i)
  if (ir == 0) { message("No individual components for ", k); next }

  is_mat <- tryCatch(
    get_block_scores(mb_fit, i, "individual"),
    error = function(e) NULL
  )
  if (is.null(is_mat)) { message("Could not get individual scores for ", k); next }

  is_df <- as.data.frame(is_mat[, 1, drop = FALSE])
  colnames(is_df) <- "IC1"
  # block scores may lack rownames; use meta_s$sample (same ordering)
  is_df$sample <- meta_s$sample
  is_df <- merge(is_df, meta_s[, c("sample",
                  intersect(c("Category", "Diagnosis"), colnames(meta_s))),
                  drop = FALSE],
                 by = "sample", all.x = TRUE)

  if ("Category" %in% colnames(is_df)) {
    p <- ggplot(is_df, aes(x = Category, y = IC1, fill = Category)) +
      geom_boxplot(alpha = 0.7) +
      geom_jitter(width = 0.12, size = 1.5, alpha = 0.5) +
      labs(title = paste("Individual component 1 —", k),
           x = "Category", y = "IC1") +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 30, hjust = 1))
    print(p)
  }
}
```

# Which joint features are statistically supported?

```{r jackstraw, fig.height=5}
js <- readRDS(file.path(data_dir, "microbiome_jackstraw_results.rds"))
print(summary(js))
plot_jackstraw(js)
```

```{r sig-features}
sig <- tryCatch(get_significant_vars(js, block = 1L, component = 1L),
                error = function(e) NULL)
if (!is.null(sig) && length(sig) > 0) {
  cat("Significant features (block 1, joint component 1):\n")
  cat(paste(head(sig, 20), collapse = "\n"), "\n")
  if (length(sig) > 20) cat("...and", length(sig) - 20, "more\n")
} else {
  message("No significant features for block 1, component 1.")
}
```

# Do clinical variables follow the joint axis?

```{r clinical-assoc}
# Categorical variables
cat_vars <- c("Category", "Diagnosis", "Sexs")
cat_vars <- intersect(cat_vars, colnames(meta_s))

if (jr >= 1 && length(cat_vars) > 0) {
  assoc_list <- lapply(cat_vars, function(v) {
    tryCatch(
      associate_components(ajive_output = mb_fit, metadata = meta_s,
                           variable = v, mode = "categorical",
                           type = "joint"),
      error = function(e) NULL
    )
  })
  assoc_list <- Filter(Negate(is.null), assoc_list)
  for (p in assoc_list) if (inherits(p, "gg")) print(p)
}
```

```{r clinical-cont}
# Continuous variables — SCFAs
cont_vars <- c("Acetate", "Propionate", "Butyrate")
cont_vars <- intersect(cont_vars, colnames(meta_s))

if (jr >= 1 && length(cont_vars) > 0) {
  assoc_cont <- lapply(cont_vars, function(v) {
    tryCatch(
      associate_components(ajive_output = mb_fit, metadata = meta_s,
                           variable = v, mode = "continuous",
                           type = "joint"),
      error = function(e) NULL
    )
  })
  assoc_cont <- Filter(Negate(is.null), assoc_cont)
  for (p in assoc_cont) if (inherits(p, "gg")) print(p)
}
```

# Does the rank survive resampling?

```{r stability, fig.height=4}
stab <- readRDS(file.path(data_dir, "microbiome_stability.rds"))
stab_summary <- tryCatch(summary(stab), error = function(e) NULL)
if (!is.null(stab_summary)) print(stab_summary)
```

# Conclusion

In this cached analysis, RaJIVE finds one joint component shared by Bacteria,
Fungi, and Viruses, plus kingdom-specific individual components. The joint axis
is the candidate ecosystem-wide signal: a host or treatment-associated pattern
that crosses kingdom boundaries. The individual components capture the other
half of the story, where each kingdom carries its own ecological structure.

The main takeaways are:

1. Bacteria, Fungi, and Viruses share a single covariation axis (`jr = 1`)
2. Jackstraw testing highlights features supporting that joint structure
3. The joint component can be screened against clinical categories
4. Individual components retain kingdom-specific biology that the joint axis
   intentionally leaves behind

# Session information

```{r sessioninfo}
sessionInfo()
```

# Appendix: Cache generation script {.appendix}

The code below recreates the cached RDS files from the source data. It needs
internet access to download from EBI FTP, so it is not evaluated during
vignette builds.

```{r cache-gen, eval=FALSE}
# ---- 0. Setup ----------------------------------------------------------------
library(rajiveplus)
library(dplyr)
library(tidyr)

out_dir <- "vignettes/data"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ---- 1. Download raw data from EBI FTP -------------------------------------
message("[1/6] Downloading raw data from EBI FTP...")
base_url <- paste0(
  "https://ftp.ebi.ac.uk/biostudies/nfs/S-BSST666/",
  "MOFA_tutorial_microbiome/")

kingdoms <- c(Bacteria = "16S_otu_abundances.csv",
              Fungi    = "ITS_otu_abundances.csv",
              Viruses  = "virome_otu_abundances.csv")
meta_file <- "sample_metadata.csv"

read_ebi <- function(file) {
  read.csv(url(paste0(base_url, file)), row.names = 1, check.names = FALSE)
}

raw_mats <- lapply(kingdoms, read_ebi)
metadata <- read_ebi(meta_file)

# CLR-normalise (add pseudo-count)
clr_norm <- function(X) {
  X <- X + 0.5
  log(X) - rowMeans(log(X))
}
raw_mats <- lapply(raw_mats, clr_norm)

# Convert to long format for storage
dt_long <- do.call(rbind, lapply(names(raw_mats), function(k) {
  m <- raw_mats[[k]]
  df <- as.data.frame(as.table(m))
  colnames(df) <- c("sample", "feature", "value")
  df$view <- k
  df[, c("sample", "feature", "view", "value")]
}))

raw <- list(dt = dt_long, metadata = metadata)
saveRDS(raw, file.path(out_dir, "microbiome_raw.rds"))
message("    -> saved microbiome_raw.rds")

# ---- 2. Preprocessing -------------------------------------------------------
message("[2/6] Preprocessing to matrix blocks...")
common_samples <- Reduce(intersect, lapply(raw_mats, rownames))
blocks <- lapply(raw_mats, function(m) m[common_samples, , drop = FALSE])
meta_s <- metadata[match(common_samples, rownames(metadata)), ]
rownames(meta_s) <- common_samples

message("    Views found: ", paste(names(blocks), collapse = ", "))
message("    Samples in common: ", length(common_samples))

pp <- list(blocks = blocks, meta = meta_s)
saveRDS(pp, file.path(out_dir, "microbiome_preprocessed.rds"))
message("    -> saved microbiome_preprocessed.rds")
message("    Block dims: ",
        paste(sapply(blocks, function(m) paste(dim(m), collapse = "x")),
              collapse = ", "))

# ---- 3. Per-block SVD for scree plots ---------------------------------------
message("[3/6] Computing per-block SVD for scree plots...")
svd_list <- lapply(blocks, function(m) {
  s <- svd(scale(m), nu = 0, nv = 0)$d
  s
})
saveRDS(svd_list, file.path(out_dir, "microbiome_svd_list.rds"))

init_ranks <- c(Bacteria = 5L, Fungi = 4L, Viruses = 3L)
message("    Initial signal ranks: ",
        paste(names(init_ranks), init_ranks, sep = "=", collapse = ", "))
message("    -> saved microbiome_svd_list.rds")

# ---- 4. Fit RaJIVE ----------------------------------------------------------
message("[4/6] Fitting Rajive() model...")
mb_fit <- Rajive(blocks, initial_signal_ranks = init_ranks)
jr <- get_joint_rank(mb_fit)
message("    Estimated joint rank: ", jr)
saveRDS(mb_fit, file.path(out_dir, "microbiome_rajive_results.rds"))
message("    -> saved microbiome_rajive_results.rds")

# ---- 5. Jackstraw -----------------------------------------------------------
message("[5/6] Running jackstraw_rajive()...")
js <- jackstraw_rajive(mb_fit, blocks = blocks, n_null = 100L)
saveRDS(js, file.path(out_dir, "microbiome_jackstraw_results.rds"))
message("    -> saved microbiome_jackstraw_results.rds")

# ---- 6. Bootstrap stability -------------------------------------------------
message("[6/6] Bootstrap rank stability (B=30)...")
stab <- assess_stability(
  ajive_output        = mb_fit,
  blocks              = blocks,
  initial_signal_ranks = init_ranks,
  target              = "joint_rank",
  B                   = 30L,
  sample_frac         = 0.8)
saveRDS(stab, file.path(out_dir, "microbiome_stability.rds"))
message("    -> saved microbiome_stability.rds")
message("All 6 cache files generated successfully.")
```
