---
title: "Native Missing-Data Analysis on Union Samples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Native Missing-Data Analysis on Union Samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette shows the native missing-data workflow for partially overlapping
multi-block studies. The important step is to align each block to the union of
sample IDs and keep absent block/sample rows as missing values, rather than
dropping samples that are not observed in every block.

```{r setup}
library(rajiveplus)
set.seed(4201)
```

## Simulate overlapping BMV-like blocks

The blocks share a latent sample trend but cover different subsets of samples,
similar to a study where not every assay was measured for every participant.

```{r simulate}
all_samples <- sprintf("sample_%02d", seq_len(30))
latent <- as.numeric(scale(seq(-1, 1, length.out = length(all_samples)) +
                             rnorm(length(all_samples), sd = 0.15)))

make_block <- function(sample_idx, n_features, prefix, noise = 0.20) {
  loadings <- seq(1.2, 0.4, length.out = n_features) *
    sample(c(-1, 1), n_features, replace = TRUE)
  x <- outer(latent[sample_idx], loadings) +
    matrix(rnorm(length(sample_idx) * n_features, sd = noise),
           nrow = length(sample_idx))
  rownames(x) <- all_samples[sample_idx]
  colnames(x) <- paste0(prefix, "_", seq_len(n_features))
  x
}

raw_blocks <- list(
  celltype = make_block(1:18, 6, "cell"),
  cytokine_6h = make_block(7:24, 5, "cy6"),
  enzyme_24h = make_block(c(1:10, 15:24), 4, "enz")
)

raw_blocks$cytokine_6h[2, 3] <- NA_real_
raw_blocks$enzyme_24h[5, 1] <- NA_real_
raw_blocks$enzyme_24h[6, ] <- NA_real_
```

## Align to the sample union

The alignment function below mirrors the preprocessing idea used by the BMV
workflow notebook: build the analysis samples by union, allocate one matrix per
block on that union, and only fill entries that were actually observed.

```{r align}
align_blocks_to_union <- function(blocks) {
  analysis_samples <- Reduce(union, lapply(blocks, rownames))
  aligned <- lapply(blocks, function(x) {
    out <- matrix(NA_real_, nrow = length(analysis_samples), ncol = ncol(x),
                  dimnames = list(analysis_samples, colnames(x)))
    idx <- match(rownames(x), analysis_samples)
    out[idx, ] <- x
    out
  })
  list(samples = analysis_samples, blocks = aligned)
}

aligned <- align_blocks_to_union(raw_blocks)
blocks_union <- aligned$blocks
mask <- lapply(blocks_union, is.finite)
complete_case <- Reduce(`&`, lapply(mask, function(m) rowSums(m) == ncol(m)))

sample_accounting <- data.frame(
  union_samples = length(aligned$samples),
  complete_case_samples = sum(complete_case),
  samples_retained_by_union = length(aligned$samples) - sum(complete_case)
)
knitr::kable(sample_accounting)
```

## Fit native missing-data RaJIVE

Native mode excludes missing cells from preprocessing, rank diagnostics, and the
observed-entry objective. Here `joint_rank = NA` triggers native candidate-rank
diagnostics and selects from the supplied candidate set.

```{r fit}
fit_union <- Rajive(
  blocks_union,
  initial_signal_ranks = c(2L, 2L, 2L),
  missing = "native",
  mask = mask,
  joint_rank = NA,
  full = FALSE,
  missing_control = rajive_missing_control(
    center = TRUE,
    scale = TRUE,
    normalize = TRUE,
    rank_candidates = 0:2
  ),
  seed = 4202
)

data.frame(
  selected_joint_rank = get_joint_rank(fit_union),
  candidate_ranks = paste(fit_union$missing$rank_diagnostics$joint_rank,
                          collapse = ", ")
)
```

## Inspect support and reconstruction provenance

The decomposition should be interpreted together with the observed support. The
estimability table records whether a reconstructed quantity is observed,
within-block estimable, cross-block joint estimable, weakly estimable, or not
identifiable.

```{r diagnostics}
info <- get_missingness_info(fit_union)
diag <- get_missing_diagnostics(fit_union)

knitr::kable(info$block_summary)
knitr::kable(diag$variance_explained)
```

```{r provenance}
missing_labels <- get_estimability(fit_union)
missing_labels <- missing_labels[!missing_labels$observed, ]
knitr::kable(utils::head(missing_labels, 12))

recon <- get_reconstructed_blocks(fit_union, type = "joint_individual",
                                  observed = "observed")
provenance <- attr(recon, "reconstruction_provenance")
knitr::kable(utils::head(provenance[!provenance$observed, ], 12))
```

The reconstructed matrices have the union sample set as their row names. Observed
entries can be restored exactly with `observed = "observed"`, while missing
entries remain explicitly labelled by the provenance table.

```{r reconstruction-shape}
vapply(recon, nrow, integer(1))
```
