---
title: "Where the Time Goes: rajiveplus vs RaJIVE"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
find_repo_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)
  repeat {
    if (file.exists(file.path(path, "DESCRIPTION")) &&
        file.exists(file.path(path, "_pkgdown.yml"))) {
      return(path)
    }
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not locate repository root from ", getwd(), call. = FALSE)
    }
    path <- parent
  }
}
repo_root <- find_repo_root()
data_dir <- file.path(repo_root, "vignettes", "data")
dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
```

## The question

Matrix decompositions ask a simple question and pay for it many times: which
directions in these blocks are shared? In Robust aJIVE, the expensive part is
the repeated robust SVD. `rajiveplus` moves that bottleneck into C++ through
RcppArmadillo, while the original
[RaJIVE](https://github.com/ericaponzi/RaJIVE) package keeps the computation in
pure R.

The useful benchmark is therefore not abstract. It is: when does that C++
implementation change the working scale of an analysis? We look along five
axes:

| Benchmark | Varies | Fixed |
|-----------|--------|-------|
| Head-to-head timing | implementation | K=2, n=100, p=500, joint_rank=2 |
| Scaling vs n | n in {50,100,200} | p=500 |
| Scaling vs p | p in {100,500,1000,2000} | n=100 |
| Parallelism (rajiveplus only) | num_cores in {1,2,4} | n=200, p=1000 |
| Peak memory | implementation | K=2, n=100, p=500 |

Two details keep the comparison honest. First, every benchmark fixes
`joint_rank = 2`, so we time the decomposition rather than stochastic rank
selection. Second, `RaJIVE` registers two parallel workers internally, so the
head-to-head runs give `rajiveplus` the same two-core budget. The direct
comparisons use `identifiability_norm = "l1"` for closer parity with original
RaJIVE; rajiveplus-only examples use the current default L2 filter.

This heavy artifact is not a package vignette. Render it only through SLURM
after the light vignette suite passes.

## Setup

This artifact does not install packages from the network. The SLURM render node
must already have the local benchmark dependencies available.

```{r setup}
required <- c("rajiveplus", "RaJIVE", "microbenchmark", "bench",
              "peakRAM", "ggplot2")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0L) {
  stop(
    "Missing local package(s) required for heavy benchmarking: ",
    paste(missing, collapse = ", "),
    ". Install them in the R4_51 environment before rendering this artifact.",
    call. = FALSE
  )
}

library(rajiveplus)
library(microbenchmark)
library(bench)
library(peakRAM)
library(ggplot2)
```

```{r parallel-setup}
doParallel::registerDoParallel(2L)
```

```{r data}
set.seed(42)
Y <- ajive.data.sim(K = 2, rankJ = 2, rankA = c(5L, 5L),
                    n = 100L, pks = c(500L, 500L), dist.type = 1)
blocks <- Y$sim_data
```

## First look

```{r first-look}
t_ru <- system.time(
  rajiveplus::Rajive(blocks, c(5L, 5L), joint_rank = 2L, full = FALSE,
                     num_cores = 2L, seed = 42L,
                     identifiability_norm = "l1")
)
t_rj <- system.time(
  RaJIVE::Rajive(blocks, c(5L, 5L), joint_rank = 2L, full = FALSE)
)
cat("rajiveplus:", round(t_ru[["elapsed"]], 2), "s\n")
cat("RaJIVE:     ", round(t_rj[["elapsed"]], 2), "s\n")
cat("Speedup:    ", round(t_rj[["elapsed"]] / t_ru[["elapsed"]], 1), "x\n")
```

## Repeated timing

```{r microbenchmark-run}
bm_mb <- microbenchmark(
  rajiveplus = rajiveplus::Rajive(
    blocks, c(5L, 5L),
    joint_rank = 2L, full = FALSE,
    num_cores = 2L, seed = 42L,
    identifiability_norm = "l1"
  ),
  RaJIVE = RaJIVE::Rajive(
    blocks, c(5L, 5L), joint_rank = 2L, full = FALSE
  ),
  times = 5L
)
saveRDS(bm_mb, file.path(data_dir, "bench_results.rds"))
```

```{r microbenchmark-table}
bm_mb <- readRDS(file.path(data_dir, "bench_results.rds"))
print(bm_mb, unit = "s", order = "median")
```

```{r microbenchmark-plot}
autoplot(bm_mb) +
  labs(title = "microbenchmark: rajiveplus vs RaJIVE",
       subtitle = "K=2, n=100, p=500, joint_rank=2, num_cores=2, 5 reps",
       x = "Time (ms)", y = NULL) +
  theme_bw()
```

## bench::mark comparison

`bench::mark` adds medians and spread. Its memory column measures R-level
allocations only, so native Armadillo allocations are addressed later with
`peakRAM`.

```{r bench-run}
bm_bench <- bench::mark(
  rajiveplus = rajiveplus::Rajive(
    blocks, c(5L, 5L),
    joint_rank = 2L, full = FALSE,
    num_cores = 2L, seed = 42L,
    identifiability_norm = "l1"
  ),
  RaJIVE = RaJIVE::Rajive(
    blocks, c(5L, 5L), joint_rank = 2L, full = FALSE
  ),
  iterations = 3L,
  check = FALSE,
  memory = FALSE
)
saveRDS(bm_bench, file.path(data_dir, "bench_results_bench.rds"))
```

```{r bench-table}
bm_bench <- readRDS(file.path(data_dir, "bench_results_bench.rds"))
knitr::kable(
  summary(bm_bench)[, c("expression", "min", "median", "mem_alloc", "n_itr")],
  caption = "bench::mark results (3 iterations)"
)
```

```{r bench-plot}
plot(bm_bench) +
  labs(title = "bench::mark: rajiveplus vs RaJIVE",
       subtitle = "R-heap allocations only; see peakRAM for total RSS") +
  theme_bw()
```

## Scaling with samples and features

```{r scaling-n}
ns <- c(50L, 100L, 200L)

scaling_n_rows <- lapply(ns, function(n_i) {
  set.seed(42)
  Y_i <- ajive.data.sim(K = 2, rankJ = 2, rankA = c(5L, 5L),
                        n = n_i, pks = c(500L, 500L), dist.type = 1)
  blks_i <- Y_i$sim_data
  t_ru <- system.time(rajiveplus::Rajive(
    blks_i, c(5L, 5L),
    joint_rank = 2L, full = FALSE,
    num_cores = 2L, seed = 42L,
    identifiability_norm = "l1"
  ))[["elapsed"]]
  t_rj <- system.time(
    RaJIVE::Rajive(blks_i, c(5L, 5L), joint_rank = 2L, full = FALSE)
  )[["elapsed"]]
  data.frame(n = n_i, rajiveplus = t_ru, RaJIVE = t_rj)
})
scaling_n_df <- do.call(rbind, scaling_n_rows)
```

```{r scaling-p}
ps <- c(100L, 500L, 1000L, 2000L)

scaling_p_rows <- lapply(ps, function(p_i) {
  set.seed(42)
  Y_i <- ajive.data.sim(K = 2, rankJ = 2, rankA = c(5L, 5L),
                        n = 100L, pks = c(p_i, p_i), dist.type = 1)
  blks_i <- Y_i$sim_data
  t_ru <- system.time(rajiveplus::Rajive(
    blks_i, c(5L, 5L),
    joint_rank = 2L, full = FALSE,
    num_cores = 2L, seed = 42L,
    identifiability_norm = "l1"
  ))[["elapsed"]]
  t_rj <- system.time(
    RaJIVE::Rajive(blks_i, c(5L, 5L), joint_rank = 2L, full = FALSE)
  )[["elapsed"]]
  data.frame(p = p_i, rajiveplus = t_ru, RaJIVE = t_rj)
})
scaling_p_df <- do.call(rbind, scaling_p_rows)

saveRDS(list(vs_n = scaling_n_df, vs_p = scaling_p_df),
        file.path(data_dir, "scaling_results.rds"))
```

```{r scaling-load}
scaling <- readRDS(file.path(data_dir, "scaling_results.rds"))
scaling_n_df <- scaling$vs_n
scaling_p_df <- scaling$vs_p
```

```{r scaling-n-plot}
df_n_long <- reshape(scaling_n_df,
                     varying = c("rajiveplus", "RaJIVE"),
                     v.names = "elapsed",
                     timevar = "package",
                     times = c("rajiveplus", "RaJIVE"),
                     direction = "long")

ggplot(df_n_long, aes(x = n, y = elapsed, colour = package, group = package)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c(rajiveplus = "#2166ac", RaJIVE = "#d6604d")) +
  labs(title = "Elapsed time vs sample size (n)",
       subtitle = "K=2, p=500, joint_rank=2, num_cores=2",
       x = "Sample size (n)", y = "Elapsed time (s)", colour = NULL) +
  theme_bw()
```

```{r scaling-p-plot}
df_p_long <- reshape(scaling_p_df,
                     varying = c("rajiveplus", "RaJIVE"),
                     v.names = "elapsed",
                     timevar = "package",
                     times = c("rajiveplus", "RaJIVE"),
                     direction = "long")

ggplot(df_p_long, aes(x = p, y = elapsed, colour = package, group = package)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_log10() +
  scale_colour_manual(values = c(rajiveplus = "#2166ac", RaJIVE = "#d6604d")) +
  labs(title = "Elapsed time vs feature count (p)",
       subtitle = "K=2, n=100, joint_rank=2, num_cores=2 (log x-axis)",
       x = "Features per block (p)", y = "Elapsed time (s)", colour = NULL) +
  theme_bw()
```

```{r speedup-table}
speedup_df <- data.frame(
  p = scaling_p_df$p,
  speedup = round(scaling_p_df$RaJIVE / scaling_p_df$rajiveplus, 1)
)
knitr::kable(speedup_df,
             col.names = c("p (features per block)", "Speedup (x)"),
             caption = "rajiveplus speedup over RaJIVE by feature count")
```

## What do extra cores buy?

```{r parallel-run}
set.seed(42)
Y_par <- ajive.data.sim(K = 2, rankJ = 2, rankA = c(5L, 5L),
                        n = 200L, pks = c(1000L, 1000L), dist.type = 1)
blks_par <- Y_par$sim_data

parallel_rows <- lapply(c(1L, 2L, 4L), function(nc) {
  t_i <- system.time(
    rajiveplus::Rajive(blks_par, c(5L, 5L), joint_rank = 2L,
                       full = FALSE, num_cores = nc, seed = 42L)
  )[["elapsed"]]
  data.frame(num_cores = nc, elapsed = t_i)
})
parallel_df <- do.call(rbind, parallel_rows)
parallel_df$speedup <- parallel_df$elapsed[1] / parallel_df$elapsed

saveRDS(parallel_df, file.path(data_dir, "parallel_results.rds"))
```

```{r parallel-table}
parallel_df <- readRDS(file.path(data_dir, "parallel_results.rds"))
knitr::kable(parallel_df, digits = 2,
             col.names = c("num_cores", "Elapsed (s)", "Speedup (x)"),
             caption = "rajiveplus parallelism scaling (n=200, p=1000)")
```

```{r parallel-plot}
p1 <- ggplot(parallel_df, aes(x = num_cores, y = elapsed)) +
  geom_line(colour = "#2166ac", linewidth = 1) +
  geom_point(size = 2.5, colour = "#2166ac") +
  labs(title = "Elapsed time vs num_cores",
       x = "num_cores", y = "Elapsed time (s)") +
  theme_bw()

p2 <- ggplot(parallel_df, aes(x = num_cores, y = speedup)) +
  geom_line(colour = "#1b7837", linewidth = 1) +
  geom_point(size = 2.5, colour = "#1b7837") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  labs(title = "Speedup vs num_cores",
       subtitle = "Dashed = ideal linear speedup",
       x = "num_cores", y = "Speedup (x)") +
  theme_bw()

if (requireNamespace("patchwork", quietly = TRUE)) {
  print(patchwork::wrap_plots(p1, p2, ncol = 2))
} else {
  print(p1)
  print(p2)
}
```

## Memory footprint

Speed is only half the story. The C++ RobustSVD allocates through Armadillo, so
R's heap counters miss part of the footprint. `peakRAM` watches the process at
the operating-system level and gives the fairer comparison.

```{r peakram-run}
mem_ru <- peakRAM::peakRAM(
  rajiveplus::Rajive(blocks, c(5L, 5L), joint_rank = 2L, full = FALSE,
                     num_cores = 2L, seed = 42L,
                     identifiability_norm = "l1")
)
mem_rj <- peakRAM::peakRAM(
  RaJIVE::Rajive(blocks, c(5L, 5L), joint_rank = 2L, full = FALSE)
)
peakram_df <- data.frame(
  package = c("rajiveplus", "RaJIVE"),
  peak_ram_mb = c(mem_ru$Peak_RAM_Used_MiB, mem_rj$Peak_RAM_Used_MiB),
  elapsed_s = c(mem_ru$Elapsed_Time_sec, mem_rj$Elapsed_Time_sec)
)
saveRDS(peakram_df, file.path(data_dir, "peakram_single.rds"))
```

```{r peakram-table}
peakram_df <- readRDS(file.path(data_dir, "peakram_single.rds"))
knitr::kable(peakram_df,
             col.names = c("Package", "Peak RAM (MiB)", "Elapsed (s)"),
             digits = 1,
             caption = "Peak RSS memory (peakRAM); includes C++ heap allocations")
```

```{r peakram-plot}
ggplot(peakram_df, aes(x = package, y = peak_ram_mb, fill = package)) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c(rajiveplus = "#2166ac", RaJIVE = "#d6604d")) +
  labs(title = "Peak RAM usage (peakRAM)",
       subtitle = "K=2, n=100, p=500, joint_rank=2, num_cores=2",
       x = NULL, y = "Peak RAM (MiB)") +
  theme_bw() +
  theme(legend.position = "none")
```

> **Why peakRAM?** `bench::mark` reports R-heap allocation. The native heap used
> by Armadillo and LAPACK is outside that accounting. `peakRAM` captures the
> process-level resident set size, which is the memory pressure the scheduler
> sees.

## Session Info

```{r session-info}
sessionInfo()
```
