---
title: "Native Missing-Data RaJIVE: Where It Scales and Where It Recovers"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r opts}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
find_repo_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)
  repeat {
    if (file.exists(file.path(path, "DESCRIPTION")) &&
        file.exists(file.path(path, "_pkgdown.yml"))) {
      return(path)
    }
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not locate repository root from ", getwd(), call. = FALSE)
    }
    path <- parent
  }
}
repo_root <- find_repo_root()
data_dir <- file.path(repo_root, "vignettes", "data")
dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)

# Smoke mode (BENCH_MISSING_SMOKE=1): one value per axis, one seed -- used to
# confirm the artifact runs end to end before committing the full SLURM grid.
SMOKE <- Sys.getenv("BENCH_MISSING_SMOKE") == "1"
```

## The question

`vignettes/native_missing_union.Rmd` shows the native missing-data workflow on
partially overlapping multi-block studies: align each block to the union of
samples, keep absent block/sample rows as missing, and fit RaJIVE over the
observed entries. The audit `audits/2026-05-15-missing-data-audit.md` found the
estimator sound but single-pass at the JIVE level, with a partial port of
missMDA's regularized EM already in place (`svd_shrinkage = "missmda"`).

This artifact measures the two things that matter for that estimator:

1. **Where does it stop scaling?** Wall time and peak memory as missingness,
   block count, cross-block overlap, and `(n, p)` size grow.
2. **Where does it stop recovering the truth?** Reconstruction RMSE at the
   unobserved cells, and whether native auto-rank recovers the true joint rank.

Every axis is crossed with `svd_shrinkage in {0, "missmda"}`, so the headline
contrast quantifies how much the partial missMDA port already buys before the
joint-level EM (audit finding F1) lands. This heavy artifact is **not** a
package vignette; render it only through SLURM.

## Setup

```{r setup}
required <- c("rajiveplus", "peakRAM", "ggplot2")
missing_pkgs <- required[!vapply(required, requireNamespace,
                                 logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0L) {
  stop("Missing local package(s) for the missing-data benchmark: ",
       paste(missing_pkgs, collapse = ", "), ".", call. = FALSE)
}
library(rajiveplus)
library(peakRAM)
library(ggplot2)

has_missmda <- requireNamespace("missMDA", quietly = TRUE)
```

```{r simulator}
# simulate_missing_benchmark(): extends vignettes/native_missing_union.Rmd.
# K blocks share a joint structure (rankJ) and each carries its own
# individual structure plus noise. A core of round(overlap_frac * n) samples
# is observed by every block; the rest are split into K block-exclusive
# slices. Each block is aligned to the n-sample union (absent samples -> NA),
# then `missing_frac` of the *observed* cells are blanked MCAR and one whole
# observed row per block is dropped (mirroring the vignette).
#
# Returns: blocks_union, mask, truth (noise-free joint+individual on the
# union), and the true joint rank.
simulate_missing_benchmark <- function(n, p, K, rankJ = 2L,
                                       overlap_frac = 0.6,
                                       missing_frac = 0.15,
                                       noise = 0.5, seed = 1L) {
  set.seed(seed)
  all_samples <- sprintf("sample_%04d", seq_len(n))
  joint_scores <- matrix(rnorm(n * rankJ), n, rankJ)

  n_core <- max(rankJ + 1L, round(overlap_frac * n))
  n_core <- min(n_core, n)
  core <- seq_len(n_core)
  rest <- setdiff(seq_len(n), core)
  slices <- if (length(rest) > 0L) {
    split(rest, cut(seq_along(rest), breaks = K, labels = FALSE))
  } else {
    replicate(K, integer(0), simplify = FALSE)
  }

  to_union <- function(mat, idx, prefix) {
    out <- matrix(NA_real_, n, ncol(mat),
                  dimnames = list(all_samples,
                                  paste0(prefix, "_", seq_len(ncol(mat)))))
    out[idx, ] <- mat
    out
  }

  blocks_union <- truth_union <- vector("list", K)
  for (k in seq_len(K)) {
    idx <- sort(c(core, slices[[k]]))
    nk <- length(idx)
    joint_loadings <- matrix(rnorm(p * rankJ, sd = 0.8), p, rankJ)
    individual_score <- as.numeric(scale(rnorm(nk)))
    individual_loadings <- rnorm(p, sd = 0.4)
    signal <- joint_scores[idx, , drop = FALSE] %*% t(joint_loadings) +
      outer(individual_score, individual_loadings)
    x <- signal + matrix(rnorm(nk * p, sd = noise), nk, p)
    prefix <- paste0("blk", k)
    blocks_union[[k]] <- to_union(x, idx, prefix)
    truth_union[[k]] <- to_union(signal, idx, prefix)
  }
  names(blocks_union) <- names(truth_union) <- paste0("blk", seq_len(K))

  mask <- lapply(blocks_union, is.finite)
  for (k in seq_len(K)) {
    obs <- which(mask[[k]])
    n_blank <- floor(missing_frac * length(obs))
    if (n_blank > 0L) mask[[k]][sample(obs, n_blank)] <- FALSE
    # Drop one whole observed row per block, picking a *distinct core* sample
    # so it stays observed in the other blocks (mirrors the vignette's
    # enzyme_24h[6, ] <- NA). Core samples are observed by every block. A
    # distinct row per block is required because the package aborts on a
    # sample observed in no block (audit finding F11) rather than dropping it.
    drop_row <- core[((k - 1L) %% length(core)) + 1L]
    mask[[k]][drop_row, ] <- FALSE
  }

  list(blocks_union = blocks_union, mask = mask, truth = truth_union,
       rankJ = rankJ)
}

# recon RMSE at the cells the model did not see but whose truth is defined.
recon_rmse <- function(fit, sim) {
  recon <- get_reconstructed_blocks(fit, type = "joint_individual",
                                    scale = "original")
  num <- den <- 0
  for (k in seq_along(recon)) {
    score_cells <- !sim$mask[[k]] & is.finite(sim$truth[[k]])
    if (!any(score_cells)) next
    num <- num + sum((recon[[k]][score_cells] - sim$truth[[k]][score_cells])^2)
    den <- den + sum(score_cells)
  }
  if (den == 0) NA_real_ else sqrt(num / den)
}

# missMDA imputeMFA reference RMSE on the same union matrix (optional).
missmda_rmse <- function(sim) {
  if (!has_missmda) return(NA_real_)
  K <- length(sim$blocks_union)
  filled <- do.call(cbind, lapply(seq_len(K), function(k) {
    x <- sim$blocks_union[[k]]
    x[!sim$mask[[k]]] <- NA_real_
    x
  }))
  grp <- vapply(sim$blocks_union, ncol, integer(1))
  imp <- tryCatch(
    missMDA::imputeMFA(as.data.frame(filled), group = grp,
                       type = rep("s", K), ncp = sim$rankJ),
    error = function(e) NULL
  )
  if (is.null(imp)) return(NA_real_)
  completed <- as.matrix(imp$completeObs)
  truth_cat <- do.call(cbind, sim$truth)
  offset <- 0
  num <- den <- 0
  for (k in seq_len(K)) {
    cols <- offset + seq_len(grp[k])
    score_cells <- !sim$mask[[k]] & is.finite(sim$truth[[k]])
    if (any(score_cells)) {
      blk_complete <- completed[, cols, drop = FALSE]
      blk_truth <- sim$truth[[k]]
      num <- num + sum((blk_complete[score_cells] - blk_truth[score_cells])^2)
      den <- den + sum(score_cells)
    }
    offset <- offset + grp[k]
  }
  if (den == 0) NA_real_ else sqrt(num / den)
}

# One grid cell: fit at fixed rank (fit cost) and at joint_rank = NA (auto-rank
# cost), record time, peak RAM, recon RMSE, rank recovery, weak support.
run_missing_cell <- function(sim, shrinkage, seed) {
  K <- length(sim$blocks_union)
  isr <- rep(max(sim$rankJ + 2L, 4L), K)
  ctrl <- rajive_missing_control(center = TRUE, scale = TRUE, normalize = TRUE,
                                 svd_shrinkage = shrinkage, rank_repeats = 1L)
  fixed_fit <- NULL
  pr <- peakRAM::peakRAM(
    fixed_fit <- Rajive(sim$blocks_union, initial_signal_ranks = isr,
                        missing = "native", mask = sim$mask,
                        joint_rank = sim$rankJ, full = FALSE,
                        missing_control = ctrl, seed = seed)
  )
  auto_fit <- Rajive(sim$blocks_union, initial_signal_ranks = isr,
                     missing = "native", mask = sim$mask,
                     joint_rank = NA, full = FALSE,
                     missing_control = ctrl, seed = seed)
  rd <- auto_fit$missing$rank_diagnostics
  selected <- get_joint_rank(auto_fit)
  weak <- if (!is.null(rd)) {
    rd$weak_support_rate[rd$joint_rank == selected][1L]
  } else NA_real_
  data.frame(
    shrinkage = if (identical(shrinkage, 0)) "none" else "missmda",
    seed = seed,
    wall_s = pr$Elapsed_Time_sec,
    peak_ram_mb = pr$Peak_RAM_Used_MiB,
    recon_rmse = recon_rmse(fixed_fit, sim),
    missmda_rmse = missmda_rmse(sim),
    selected_rank = selected,
    true_rank = sim$rankJ,
    rank_correct = selected == sim$rankJ,
    weak_support_rate = weak,
    stringsAsFactors = FALSE
  )
}

# Sweep one axis: `values` substituted into the baseline, crossed with
# {0, "missmda"} shrinkage and `seeds`.
run_sweep <- function(axis, values, baseline, seeds) {
  rows <- list()
  for (v in values) {
    params <- baseline
    params[[axis]] <- v
    for (sh in list(0, "missmda")) {
      for (s in seeds) {
        sim <- do.call(simulate_missing_benchmark,
                       c(params, list(seed = s)))
        cell <- run_missing_cell(sim, sh, s)
        cell[[axis]] <- v
        rows[[length(rows) + 1L]] <- cell
      }
    }
  }
  out <- do.call(rbind, rows)
  out[, c(axis, setdiff(names(out), axis))]
}
```

```{r grid}
baseline <- list(n = 120L, p = 100L, K = 3L, rankJ = 2L,
                 overlap_frac = 0.6, missing_frac = 0.15, noise = 0.5)
seeds <- if (SMOKE) 1L else 1:3

grids <- if (SMOKE) {
  list(missing_frac = 0.30, K = 3L, overlap_frac = 0.6, n = 120L)
} else {
  list(
    missing_frac = c(0.05, 0.15, 0.30, 0.50),
    K            = c(2L, 3L, 5L),
    overlap_frac = c(0.3, 0.6, 0.9),
    n            = c(60L, 120L, 240L, 480L)
  )
}
# the `n` sweep scales p with n so both grow together
size_baseline <- baseline
```

## Sweep 1 — missingness fraction

```{r sweep-missingness}
bench_missingness <- run_sweep("missing_frac", grids$missing_frac,
                               baseline, seeds)
saveRDS(bench_missingness,
        file.path(data_dir, "bench_missing_missingness.rds"))
knitr::kable(bench_missingness, digits = 3,
             caption = "Native missing-data fit vs missingness fraction")
```

## Sweep 2 — number of blocks

```{r sweep-nblocks}
bench_nblocks <- run_sweep("K", grids$K, baseline, seeds)
saveRDS(bench_nblocks, file.path(data_dir, "bench_missing_nblocks.rds"))
knitr::kable(bench_nblocks, digits = 3,
             caption = "Native missing-data fit vs number of blocks")
```

## Sweep 3 — cross-block overlap

```{r sweep-overlap}
bench_overlap <- run_sweep("overlap_frac", grids$overlap_frac, baseline, seeds)
saveRDS(bench_overlap, file.path(data_dir, "bench_missing_overlap.rds"))
knitr::kable(bench_overlap, digits = 3,
             caption = "Native missing-data fit vs cross-block overlap")
```

## Sweep 4 — problem size (n, with p scaled alongside)

```{r sweep-size}
size_rows <- list()
for (n_i in grids$n) {
  params <- size_baseline
  params$n <- n_i
  params$p <- as.integer(round(n_i * 0.8))
  for (sh in list(0, "missmda")) {
    for (s in seeds) {
      sim <- do.call(simulate_missing_benchmark, c(params, list(seed = s)))
      cell <- run_missing_cell(sim, sh, s)
      cell$n <- n_i
      cell$p <- params$p
      size_rows[[length(size_rows) + 1L]] <- cell
    }
  }
}
bench_size <- do.call(rbind, size_rows)
bench_size <- bench_size[, c("n", "p",
                             setdiff(names(bench_size), c("n", "p")))]
saveRDS(bench_size, file.path(data_dir, "bench_missing_size.rds"))
knitr::kable(bench_size, digits = 3,
             caption = "Native missing-data fit vs problem size")
```

## Summary plots

```{r plot-rmse-missingness}
agg <- function(df, axis) {
  stats::aggregate(
    cbind(wall_s, peak_ram_mb, recon_rmse, rank_correct, weak_support_rate) ~
      get(axis) + shrinkage,
    data = df, FUN = function(x) mean(x, na.rm = TRUE)
  ) |> setNames(c(axis, "shrinkage", "wall_s", "peak_ram_mb",
                  "recon_rmse", "rank_recovery", "weak_support_rate"))
}

ggplot(agg(bench_missingness, "missing_frac"),
       aes(missing_frac, recon_rmse, colour = shrinkage)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  labs(title = "Reconstruction RMSE vs missingness fraction",
       subtitle = "K=3, n=120, p=100, overlap=0.6; lower is better",
       x = "Missingness fraction", y = "RMSE at unobserved cells") +
  theme_bw()
```

```{r plot-rank-overlap}
ggplot(agg(bench_overlap, "overlap_frac"),
       aes(overlap_frac, rank_recovery, colour = shrinkage)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  ylim(0, 1) +
  labs(title = "Auto-rank recovery vs cross-block overlap",
       subtitle = "Fraction of fits selecting the true joint rank",
       x = "Cross-block overlap fraction", y = "Rank recovery rate") +
  theme_bw()
```

```{r plot-time-size}
ggplot(agg(bench_size, "n"),
       aes(n, wall_s, colour = shrinkage)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  labs(title = "Wall time vs problem size",
       subtitle = "p scaled as 0.8 * n; fixed-rank fit",
       x = "Sample size n", y = "Elapsed time (s)") +
  theme_bw()
```

```{r missmda-note}
if (has_missmda) {
  cat("missMDA::imputeMFA reference RMSE is recorded in the `missmda_rmse`",
      "column of every cached data frame for direct comparison.\n")
} else {
  cat("missMDA is not installed; `missmda_rmse` is NA. Install missMDA on the",
      "render node to enable the regularized-EM reference comparison.\n")
}
```

## Session info

```{r session-info}
sessionInfo()
```
