---
title: "BMV Native Missing-Data Union Analysis"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This local analysis artifact exercises native missing-data RaJIVE on the BMV
preprocessing cache. It is intentionally kept outside package vignettes because
it depends on local HPC data and scratch cache paths.

```{r setup}
repo_root <- Sys.getenv("RAJIVEPLUS_REPO_ROOT", unset = getwd())
repo_root <- normalizePath(repo_root, winslash = "/", mustWork = TRUE)

source(file.path(repo_root, "scratch", "bmv_workflow", "config.R"), chdir = TRUE)
config <- define_bmv_publication_config(repo_root = repo_root)
source_bmv_workflow_helpers(config)

library(rajiveplus)
```

## Load union-sample preprocessing

```{r load-preprocess}
preprocess_path <- file.path(config$paths$objects, "preprocess_objects.rds")
if (!file.exists(preprocess_path)) {
  stop("BMV preprocess cache is missing: ", preprocess_path,
       "\nRun the BMV preprocess stage before rendering this artifact.",
       call. = FALSE)
}

preprocess <- readRDS(preprocess_path)
global_union_samples <- length(preprocess$analysis_samples)
```

The BMV preprocessing cache contains finite `scaled_blocks` for the older
complete-data fit path, plus `aligned_blocks` that preserve missing assay rows as
`NA`. For native missing-data fitting, use the scaled values but restore the
missingness pattern from `aligned_blocks`.

```{r restore-native-missing}
restore_native_missing_blocks <- function(preprocess) {
  block_names <- names(preprocess$scaled_blocks)
  out <- lapply(block_names, function(nm) {
    scaled <- preprocess$scaled_blocks[[nm]]
    aligned <- preprocess$aligned_blocks[[nm]]
    missing_features <- setdiff(colnames(scaled), colnames(aligned))
    if (length(missing_features) > 0L) {
      stop("Scaled block features are absent from aligned block for ", nm,
           ": ", paste(missing_features, collapse = ", "), call. = FALSE)
    }
    template <- aligned[rownames(scaled), colnames(scaled), drop = FALSE]
    native <- scaled
    native[!is.finite(template)] <- NA_real_
    native
  })
  names(out) <- block_names
  out
}

native_blocks_all <- restore_native_missing_blocks(preprocess)
```

## Select the render scope

By default, this artifact fits a small subset so it can be used as a fast smoke
test. Set `RUN_FULL_BMV_NATIVE=1` in the SLURM environment to fit all blocks and
features.

```{r subset}
run_full <- identical(Sys.getenv("RUN_FULL_BMV_NATIVE"), "1")

native_blocks <- native_blocks_all
if (!run_full) {
  native_blocks <- native_blocks[seq_len(min(3L, length(native_blocks)))]
  native_blocks <- lapply(native_blocks, function(x) {
    x[, seq_len(min(12L, ncol(x))), drop = FALSE]
  })
}

observed_in_any_selected_block <- Reduce(`|`, lapply(native_blocks, function(x) {
  rowSums(is.finite(x)) > 0L
}))
fit_blocks <- lapply(native_blocks, function(x) {
  x[observed_in_any_selected_block, , drop = FALSE]
})

feature_supported <- lapply(fit_blocks, function(x) colSums(is.finite(x)) > 0L)
fit_blocks <- Map(function(x, keep) x[, keep, drop = FALSE],
                  fit_blocks, feature_supported)
mask <- lapply(fit_blocks, is.finite)

complete_case <- Reduce(`&`, lapply(mask, function(m) rowSums(m) == ncol(m)))
block_support <- data.frame(
  block = names(fit_blocks),
  n_samples = vapply(fit_blocks, nrow, integer(1L)),
  n_features = vapply(fit_blocks, ncol, integer(1L)),
  whole_missing_sample_rows = vapply(mask, function(m) sum(rowSums(m) == 0L),
                                     integer(1L)),
  observed_fraction = vapply(mask, mean, numeric(1L)),
  stringsAsFactors = FALSE
)

sample_accounting <- data.frame(
  global_union_samples = global_union_samples,
  selected_fit_samples = nrow(fit_blocks[[1L]]),
  complete_case_samples = sum(complete_case),
  run_full_bmv_native = run_full,
  stringsAsFactors = FALSE
)

knitr::kable(sample_accounting)
knitr::kable(block_support)
```

## Fit native missing-data RaJIVE

```{r fit}
initial_signal_ranks <- vapply(fit_blocks, function(x) {
  as.integer(min(if (run_full) 6L else 2L, nrow(x), ncol(x)))
}, integer(1L))

candidate_max <- min(initial_signal_ranks, if (run_full) 4L else 2L)
rank_candidates <- seq.int(0L, candidate_max)

fit_native <- Rajive(
  fit_blocks,
  initial_signal_ranks = initial_signal_ranks,
  missing = "native",
  mask = mask,
  joint_rank = NA,
  full = FALSE,
  missing_control = rajive_missing_control(
    center = FALSE,
    scale = FALSE,
    normalize = FALSE,
    rank_candidates = rank_candidates
  ),
  seed = 20260514L
)

rank_summary <- data.frame(
  selected_joint_rank = get_joint_rank(fit_native),
  candidate_ranks = paste(fit_native$missing$rank_diagnostics$joint_rank,
                          collapse = ", "),
  stringsAsFactors = FALSE
)
knitr::kable(rank_summary)
```

## Diagnostics and provenance

```{r diagnostics}
diag <- get_missing_diagnostics(fit_native)
knitr::kable(diag$variance_explained)
knitr::kable(utils::head(fit_native$missing$rank_diagnostics, 10L))
```

```{r provenance}
recon <- get_reconstructed_blocks(fit_native, type = "joint_individual",
                                  observed = "observed")
provenance <- attr(recon, "reconstruction_provenance")
missing_provenance <- provenance[!provenance$observed, , drop = FALSE]

knitr::kable(utils::head(missing_provenance, 20L))
```

```{r reconstruction-dimensions}
reconstruction_dimensions <- data.frame(
  block = names(recon),
  n_samples = vapply(recon, nrow, integer(1L)),
  n_features = vapply(recon, ncol, integer(1L)),
  stringsAsFactors = FALSE
)
knitr::kable(reconstruction_dimensions)
```
